---
execute:
  freeze: auto
---

# Methods

```{r setup, include=FALSE, eval=TRUE}
## Setting up
 pacman::p_load(galah, arrow, here, tidyverse, dplyr, janitor, ozmaps, sf, ggplot2, skimr, sp, purrr, ConR, dplyr, DT, zoomerjoin, gt)

# Load helper functions
source(here("R/utils.R"))

## Read in parquet of raw download
myg_spiders_all <- open_dataset(get_latest_download(path_to_galah_data = paste0(here("data/galah/"), "/")))

# Load cleaned data
myg_spiders_cleaned <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[2])

# Collect data for overview and cleaning
## Subset columns we want for ALA data overview
myg_spiders <- myg_spiders_all |> 
  dplyr::select(scientificName, decimalLatitude:collectionCode) |> 
  collect()

# AFD parquet used for Data cleaning
afd_myg <- read_csv(here("data/afd_myg_08-01-2026.csv"))
```

## Data retrieval

::: callout-note
We downloaded data for this report on the `r get_latest_download(paste0(here("data/galah/"), "/")) |> stringr::str_extract(pattern = regex("[0-9]{4}\\-[0-9]{2}\\-[0-9]{2}"))` and the raw download contained `r nrow(myg_spiders_all)` records
:::

::: {.callout-tip title="Source code"}
The data retrieval workflow can be found [here](https://github.com/invertsau/inverthotspot/blob/main/R/01_ala_data_retrieval_mygalomorph.Rmd)
:::

We used data from [Atlas of Living Australia (ALA)]() for this study. We downloaded occurrence records using the `galah` [R package]() using the following criteria:

1.  Found in Australian mainland or Tasmania.
2.  Identified to a taxon rank of species.
3.  Basis of record of either:
    i)  Preserved specimen
    ii) Material sample
    iii) Machine observation
    iv) Human observation
4.  Coordinate uncertainty of less than 1000 meters or has a value of NA (citizen science records or human observations are typically entered as NA)

We also used ALA's data quality assertions to further refine our download. We excluded occurrence records using the following criteria:

1.  Coordinates are equal to 0
2.  Coordinates are presumed swapped e.g. when latitude is entered as longitude
3.  Latitude and longitude values are presumed negated
4.  Coordinates are out of range
5.  Taxon excluded by the ALA
6.  Taxon considered as a questionable species

See @sec-assertions to learn more about which assertions we used for this project.

## Data overview

::: callout-note
The full data overview report can be found [here](https://posit.cloud/content/7700631)
:::

::: {.callout-tip title="Source code"}
The data overview workflow can be found [here](https://github.com/invertsau/inverthotspot/blob/main/R/02_ala_data_overview_mygalomorph.Rmd)
:::

After the initial data retrieval the data were summarised to provide an overview of the number of records, species, and families represented, as well as the broad distribution of these across Australian states. The basis of records and data quality assertions were also investigated. This provided a broad overview of data quality and facilitated review by taxonomic experts.

### Species

```{r, echo=FALSE}
# Total number of species
total_myg <- myg_spiders |> 
  filter(!is.na(species)) |> 
  distinct(species) |> 
  nrow()
```

There are **`r total_myg` species** in total in the original download.

The below barplot shows the distribution of the number of records across species.

```{r, echo = FALSE}
# number of records for each species
myg_records <- myg_spiders |>
  filter(!is.na(species)) |>
  group_by(species) |>
  tabyl(species) |>
  arrange(desc(n))

myg_records_df <- as.data.frame(myg_records)
barplot(myg_records_df$n, xlab = "species", ylab = "records")

```

The below table shows the number of records of each species.

```{r, echo = FALSE}
myg_records |> 
  mutate(percent = percent*100) |>
  datatable() |>
  formatRound(columns="percent", digits=2)
```

### Family

Barplot showing the number of species in each family:

```{r, echo = FALSE, message=FALSE}
# N species, total number of species (bar chart or alike, or tabset)
# N species per family
# number of species in each family
myg_fams <- myg_spiders |>
  filter(!is.na(species)) |>
  group_by(species) |>
  #filter(n() > 2) |>
  group_by(family, species) |>
  summarise(n = n()) |>
  tabyl(family) |>
  arrange(desc(n))

as.data.frame(myg_fams)|>
  mutate(family = factor(family, levels = family))|>
  ggplot(aes(x = family, y = n)) +
  geom_bar(stat = "identity") +
  labs(y = "number of species") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(colour = "black"))
#which families are not represented?

```

### Taxonomic overview by state

```{r, include=FALSE}
# number of records by state, for species with at least three overall records
myg_records_by_state <- myg_spiders |>
  filter(!is.na(species)) |>
  filter(stateProvince %in% c("Queensland", "Western Australia", "New South Wales", "South Australia", 
                              "Victoria", "Northern Territory", "Tasmania", "Australian Capital Territory")) |>
  #group_by(species) |>
  #filter(n() > 2) |>
  group_by(stateProvince, species) |>
  summarise(n = n()) |>
  arrange(stateProvince, desc(n))

# Top 3 most recorded species in each state
tab <- slice_head(myg_records_by_state, n = 3) |>
  print(n = Inf) |> 
  group_by(stateProvince)
```

```{r, echo = FALSE}
 tab |> gt(row_group_as_column = TRUE)
```

```{r, echo = FALSE, message=FALSE}
#| fig-cap: 
#|   - "Barplot showing the number of species and families by state"
# number of species by state, for species with at least three overall records
myg_spp_by_state <- myg_records_by_state |> 
  tabyl(stateProvince) |>
  arrange(desc(n))

# number of families by state
myg_fam_by_state <- myg_spiders |>
  filter(!is.na(species)) |>
  filter(stateProvince %in% c("Queensland", "Western Australia", "New South Wales", "South Australia", 
                              "Victoria", "Northern Territory", "Tasmania", "Australian Capital Territory")) |>
  group_by(species) |>
 # filter(n() > 2) |>
  group_by(stateProvince, family) |>
  summarise(n()) |>
  tabyl(stateProvince) |>
  arrange(desc(n))

#plotting numbers of species and families by state
myg_spp_fam_by_state <- left_join(myg_spp_by_state, myg_fam_by_state, by = "stateProvince")

scale_right <- max(myg_spp_fam_by_state$n.y)/max(myg_spp_fam_by_state$n.x)

as.data.frame(myg_spp_fam_by_state)|>
  mutate(stateProvince = factor(stateProvince, levels = stateProvince))|>
  ggplot(aes(x = stateProvince)) +
  geom_bar(stat = "identity", aes(y = n.x, fill = "species")) +
  geom_bar(stat = "identity", aes(y = n.y/scale_right, fill = "families"), colour = "black", alpha = 0.5) +
  scale_y_continuous(sec.axis = sec_axis(~.*scale_right, name = "number of families")) + 
  scale_fill_manual(values = c("grey30", "lightgreen")) +
  labs(y = "number of species") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(colour = "black"))
```

```{r, include = FALSE}
# map of Australia with states
aus <- st_transform(ozmaps::ozmap_states, 4326)

# checking for differences between state names
setdiff(aus$NAME, myg_spp_by_state$stateProvince)
setdiff(myg_spp_by_state$stateProvince, aus$NAME)

# checking distribution of data
hist(myg_spp_by_state$n)
hist(log(myg_spp_by_state$n))

# joining map with species count data
myg_spp_aus <- myg_spp_by_state |>
  full_join(y = aus, by = c("stateProvince" = "NAME")) |>
  st_as_sf()
```

```{r, echo=FALSE}
#| fig-cap: 
#|   - "Map showing the number of species by state"
# plotting number of species by state
ggplot() + 
  geom_sf(data = myg_spp_aus,
          aes(fill = n),
          colour = "black") +
  scale_fill_distiller(name = "state_spp",
                       type = "seq",
                       palette = "YlOrBr",
                       direction = 1,
                       guide = guide_colourbar(direction = "horizontal",
                                               label.position = "bottom",
                                               title = NULL)) +
  annotate("text", 
           x = 133, 
           y = -45.5, 
           label = "No. of species",
           size = 6) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(12, 'mm'))  
```

```{r, include = FALSE}

# map of Australia with states
aus <- st_transform(ozmaps::ozmap_states, 4326)

# checking for differences between state names
setdiff(aus$NAME, myg_fam_by_state$stateProvince)
setdiff(myg_fam_by_state$stateProvince, aus$NAME)

# checking distribution of data
hist(myg_fam_by_state$n)
```

```{r, echo=FALSE}
#| fig-cap: 
#|   - "Map showing the number of families by state"
# joining map with species count data
myg_fam_aus <- myg_fam_by_state |>
  full_join(y = aus, by = c("stateProvince" = "NAME")) |>
  st_as_sf()

# plotting number of species by state
ggplot() + 
  geom_sf(data = myg_fam_aus,
          aes(fill = n),
          colour = "black") +
  scale_fill_distiller(name = "state_fam",
                       type = "seq",
                       palette = "YlOrBr",
                       direction = 1,
                       guide = guide_colourbar(direction = "horizontal",
                                               label.position = "bottom",
                                               title = NULL)) +
  annotate("text", 
           x = 133, 
           y = -45.5, 
           label = "No. of families",
           size = 6) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(12, 'mm'))  
```

### [Basis of record](https://gbif.github.io/parsers/apidocs/org/gbif/api/vocabulary/BasisOfRecord.html)

Total counts of basis of record types.

```{r, echo = FALSE}
# Counts of basis of record types. "OCCURRENCE" and "OBSERVATION" types were excluded as they could not be traced back to a source.
myg_spiders |>
  filter(!is.na(species),
         !basisOfRecord %in% c("OCCURRENCE", "OBSERVATION")) |>
  tabyl(basisOfRecord) |>
  mutate(percent = percent*100) |>
  arrange(desc(n)) |>
  adorn_rounding(digits = 2, columns = percent) |> gt()
```

`PRESERVED_SPECIMEN`: An occurrence record describing a preserved specimen.\
`HUMAN_OBSERVATION`: An occurrence record describing an observation made by one or more people.\
`MATERIAL_SAMPLE`: An occurrence record based on samples taken from other specimens or the environment.\
`MACHINE_OBSERVATION`: An occurrence record describing an observation made by a machine.

### Assertions {#sec-assertions}

Various tests are run on occurrence data in the Atlas of Living Australia, resulting in assertions about the content and quality of the data. These assertions help users gauge whether data is fit for their purposes and allow for easy data filtering. Assertions are logical variables (`TRUE/FALSE`) and take the value TRUE when they apply to the associated occurrence record. Descriptions of Assertions can be found at [here](https://github.com/AtlasOfLivingAustralia/ala-dataquality/wiki.)

Here, we have focused on spatial and taxonomic assertions because accurate identification of taxa and their spatial distributions is imperative to the calculation of endemism metrics and subsequent mapping.

#### Spatial assertions

We investigated a range of spatial assertions and deemed the following safe to bypass when refining the data download:

-   `COORDINATE_UNCERTAINTY_METERS_INVALID`\
    We performed visual checks and flagged records did not seem to be out of species' range. They were mostly cases where `coordinateUncertaintyinMeters` was NA. We know iNaturalist records input NA for `coordinateUncertaintyinMeters`, therefore in refining the data download we will include records where `coordinateUncertaintyinMeters` is NA or less than 1000 m. We will remove flagged values for a sensitivity analysis later on.
-   `COORDINATE_ROUNDED`\
    The original coordinates were rounded to six decimals (\~ 1 m precision) to simplify processing. The level of precision lost will not affect the endemism analysis.

The following were used to refine the data download as they indicate coordinates outside of the given country (records flagged as TRUE were excluded):

-   `COORDINATE_OUT_OF_RANGE`
-   `PRESUMED_NEGATED_LONGITUDE`
-   `PRESUMED_NEGATED_LATITUDE`
-   `PRESUMED_SWAPPED_COORDINATE`
-   `ZERO_COORDINATE`

#### Taxonomic assertions

The following assertions were used to identify taxonomic discrepancies in the data and allow for review by taxonomic experts:

`TAXON_MATCH_FUZZY` - is flagged when the supplied scientific name (`raw_scientificName`) does not exactly match the taxonomic backbone of the Atlas.

## Data cleaning

::: callout-note
The full data cleaning report can be found [here](https://posit.cloud/content/7700644)
:::

::: {.callout-tip title="Source code"}
The data cleaning workflow can be found [here](https://github.com/invertsau/inverthotspot/blob/main/R/03_ala_data_clean_mygalomorph.Rmd)
:::

The following criterion were used to identify and remove records for the biodiversity analyses:

-   Taxonomic errors, invalid species names or synonyms
-   Species introduced to Australia
-   Marine taxa
-   Records with geographic errors
-   Any subspecies level identifications were reclassified to species level
-   Duplicates

### Taxonomic errors

Here we clean the data for taxonomic errors, such as misspelt and outdated names. We verified that all records have species level taxonomic data.

The [Australian Faunal Directory (AFD)](https://biodiversity.org.au/afd/home) is an online catalogue of taxonomic and biological information on all animal species known to occur within Australia and its territories. We used AFD data to ensure all species names in the ALA dataset are valid.

```{r, include=FALSE}
# how many species in ALA data matched species in the AFD?
afd_myg_valid <- afd_myg |> filter(!is.na(SPECIES),
                                   NAME_TYPE == "Valid Name")

AFD_matched <- myg_spiders |>
  filter(scientificName %in% word(afd_myg_valid$SCIENTIFIC_NAME, 1,2, sep = " ")) |>
  pull(scientificName) |>
  unique() |>
  length()

AFD_unmatched <- myg_spiders |>
  filter(!scientificName %in% word(afd_myg_valid$SCIENTIFIC_NAME, 1,2, sep = " ")) |>
  pull(scientificName) |>
  unique()

# Compute reporting stats
length(AFD_unmatched)
unique(myg_spiders$scientificName) |> length()
```

We identified `r length(AFD_unmatched)`/`r unique(myg_spiders$scientificName) |> length()` species in the ALA that did not match with names in the AFD (`r ( length(AFD_unmatched) / unique(myg_spiders$scientificName) |> length() )*100 |> round(2)`%). This species, *Teyl* sp. (MYG693), although not formally named, is listed as critically endangered in Western Australia, and has been retained in the analysis.

```{r, include=FALSE}
 #After consulting experts and checking the [World Spider Catalogue (WSC)](https://wsc.nmbe.ch/) it was determined that *Selenocosmia crassipes* is the correct name, and all *Phlogius crassipes* records were changed in the ALA dataset.
myg_spiders_no_unmatched <- myg_spiders |> 
  mutate(case_when(scientificName == "Phlogius crassipes" ~ "Selenocosmia crassipes",
                   .default = scientificName),
         case_when(species == "Phlogius crassipes" ~ "Selenocosmia crassipes",
                   .default = species))
```

### Introduced Species

Any species which are not native to Australia were removed from our dataset. To identify all introduced species in the dataset, we used species lists from the [World Spider Catalogue (WSC)](https://wsc.nmbe.ch/) and the [Global Register of Invasive and Introduced Species (GRIIS)]((http://griis.org/)).

The (WSC) is a comprehensive online database of spiders from around the world, with detailed taxonomic information, distribution maps, references and images.

```{r, include=FALSE}
wsc <- read_csv(here("data/wsc/species_export_20250814.csv"))

introduced_taxa <- wsc |> 
  filter(str_detect(distribution, pattern = "Introduced to.*Australia")) 

introduced_taxa_wfullname <- introduced_taxa |> 
  mutate(scientific_name = paste(genus, species)) 
```

::: callout-important
Introduced species in the WSC matched with `r length(which(myg_spiders_no_unmatched$scientificName %in% introduced_taxa_wfullname$scientific_name))` records in our data.
:::

(GRIIS) is a project by the IUCN SSC Invasive Species Specialist Group to compile annotated and verified country-wise inventories of introduced and invasive species.

```{r, include=FALSE}
# Join distrubtion.txt and taxon-edit.txt
griis_distrib <- read.delim(here("data/griis-australia-v1.6/distribution.txt"))
griis_taxon <- read.delim(here("data/griis-australia-v1.6/taxon-edited.txt"))
# check for missing values
setdiff(griis_distrib$id, griis_taxon$id) |> length() # 24 missing
setdiff(griis_taxon$id, griis_distrib$id) # none missing
# join lists
griis_all <- left_join(griis_distrib, griis_taxon, by = "id")  # Apply left_join dplyr function 

# filter by class = arachnida
str_subset(griis_all$class, "Arachnida") # there are no introduced arachnids in this list
```

::: callout-important
There are no arachnids in the Australian GRIIS list
:::

::: callout-note
For the purpose of this project, we conclude there are no introduced spiders in our dataset.
:::

### Marine Species

```{r, include=FALSE}
wormsOutput <- readRDS(here("output/2024-10-17_worrms_myg_wm_records_taxamatch.rds"))
```

The World Register of Marine Species (WoRMS) was used to identify and remove marine species from the data. WoRMS provides a authoritative and comprehensive list of names of marine organisms, including currently valid and alternative names.

Using the [`worrms` R package](https://docs.ropensci.org/worrms/), we supplied our list of `r unique(myg_spiders_no_unmatched$scientificName) |> length()` species to the `wm_records_taxamatch()` function to check whether any our taxa are classified as marine.

::: callout-note
In the case of this project, we found `r length(which(wormsOutput$match_type %in% "exact"))` taxa that matched with marine species in the WoRMS database.
:::

### Geographic Errors

```{r, include=FALSE}
aus <- st_transform(ozmaps::ozmap_country, 4326)

base_map <- ggplot() + 
  geom_sf(data = aus, fill = NA) + 
  theme_minimal()
```

Biodiversity data, especially citizen science data can have various geographic imperfections. For example, occurrence records may include data from institutions such as botanic gardens or zoos which may not be reflective of a specie's natural range. There may be data entry errors where the default location of a taxa is assumed to be the center of particular region. For these reasons, we used the R package [`CoordinateCleaner`](https://docs.ropensci.org/CoordinateCleaner/index.html) to investigate potential errors in our dataset.

We used the `clean_coordinates()` function and tested whether any of our coordinates were the following:

-   General coordinate validity
-   Country and province centroids
-   Capital coordinates
-   Coordinates of biodiversity institutions
-   Spatial outliers
-   Temporal outliers
-   Assigned to the location of the GBIF headquarters
-   Located in Urban areas
-   Located in the sea
-   Duplicated values
-   Plain zeros

Table of the number of observations flagged by CoordinateCleaner's testing suite:

```{r, echo=FALSE, eval=TRUE}
cc_tests <- readRDS(here("output/CoordinateCleaner_flagged.rds"))

cc_summary <- CoordinateCleaner:::summary.spatialvalid(cc_tests)

tibble(Test = names(cc_summary), 
         `Number of Observations` = as.vector(cc_summary)) |> 
  gt()
```

CoordinateCleaner, identified a large number of duplicates which we have a separate workflow in @sec-dups so we decided to ignore the duplicates flagged by CoordinateCleaner.

After some consideration:

-   `r cc_summary[".cap"] |> as.vector()` records excluded due to being near a capital vicinity (`r (( cc_summary[".cap"] |> as.vector() / nrow(myg_spiders)) *100 ) |> round(2)`%)
-   `r cc_summary[".inst"] |> as.vector()` records excluded due to being in the vicinity of biodiversity institutions (`r (( cc_summary[".inst"] |> as.vector() / nrow(myg_spiders)) *100 )|> round(2)`%).
-   `r cc_summary[".sea"] |> as.vector()` records excluded due to being in sea (`r (( cc_summary[".sea"] |> as.vector() / nrow(myg_spiders)) *100 )|> round(2)`%).

```{r, include=FALSE}
myg_spiders_geo_clean <- cc_tests |> 
  filter(!.cap == FALSE,
         !.inst == FALSE, 
         !.sea == FALSE)
```

::: callout-note
After removing these geographic errors, we have `r nrow(myg_spiders_geo_clean)` observations after the exclusion, `r unique(myg_spiders_geo_clean$scientificName) |> length()` species.
:::

### Duplicate Records {#sec-dups}

Duplicate records can affect statistical analyses by pseudoreplication (increasing sample size but not adding information). We noticed that some records that had the *same* coordinate values for a large number of *different* species. These records were typically from museum record providers, we suspect this was due to the sampling method e.g. pitfall traps. For this reason, we decided to remove records if they fulfiled the following critiera:

-   the record was duplicated at the species level
-   had same exact coordinates
-   had same time and date of collection

```{r, include=FALSE}
deduplicated_myg_spiders <- myg_spiders_geo_clean |> 
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) |> 
   distinct(scientificName, decimalLatitude, decimalLongitude, eventDate, .keep_all = TRUE)

dups <- myg_spiders_geo_clean |> 
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) |> 
   group_by(scientificName, decimalLatitude, decimalLongitude, eventDate) |> 
   filter(!n() == 1) 
```

::: callout-important
Using this method, we excluded `r nrow(myg_spiders_geo_clean) - nrow(deduplicated_myg_spiders)` records from the dataset (`r (nrow(myg_spiders_geo_clean) - nrow(deduplicated_myg_spiders))/nrow(myg_spiders_geo_clean) *100 |> round(2)`%).
:::

::: callout-note
After deduplication, we have `r nrow(deduplicated_myg_spiders)` observations and `r unique(deduplicated_myg_spiders$scientificName) |> length()` species.
:::

### Citizen science data

While citizen science is an increasingly important source of data that informs research and management, it is impacted by high sampling bias and other data quality issues such as improper species identification. As such, we decided to conduct the analysis using multiple datasets: preserved specimen-only data (basis of record = `"PRESERVED_SPECIMEN"`), citizen science data (basis of record = `"HUMAN_OBSERVATION"`), and the full (cleaned) dataset. 

```{r, include = FALSE}
expert_records <- myg_spiders_cleaned |> 
  filter(!basisOfRecord == "HUMAN_OBSERVATION")

citsci <- myg_spiders_cleaned |> 
  filter(basisOfRecord == "HUMAN_OBSERVATION")
```

::: callout-important
There were `r nrow(citsci)` records that were flagged with basis of record = `"HUMAN_OBSERVATION"`. This resulted in an exclusion of `r (( nrow(citsci) / nrow(myg_spiders) )*100) |> round(2)`% of the original data) in the "preserved specimen only" dataset.
:::

## Generating species distributions

::: {.callout-tip title="Source code"}
The alpha-hull workflow can be found [here](https://github.com/invertsau/inverthotspot/blob/main/R/04_alphahulls_mygalomorph.Rmd)
:::

Spatial data for invertebrates are typically sparse, often resulting in exclusion of entire species when using advanced spatial modelling methods such as species distribution models.

One way to retain as many species as possible in a spatial analysis is to represent species range with alpha-hulls, a form of spatial polygon. An alpha-hull can be computed using a minimum of three data points, which allows the inclusion of data-deficient invertebrates in spatial analyses. 

Using the cleaned data, we applied the `EOO.computing()` function from R package [`ConR`](https://github.com/gdauby/ConR) to quantify alpha-hulls for all species that had at least 3 records (n = `r # myg_spiders_cleaned$species |> unique() |> length()`). We used an alpha value of 2 and a buffer size of 0.000001 (which approximates to \~100m). The alpha-hull workflow is laid out in [this document](https://github.com/invertsau/inverthotspot/blob/main/R/04_alphahulls_mygalomorph.Rmd).

Alpha-hulls were converted into `SpatialPolgyon` objects using the [`sf`](https://cran.r-project.org/web/packages/sf/index.html) R package. Species with less than three records were then added as `MULTIPOINT` objects and the data exported for biodiversity analysis.

```{r, include=FALSE}

pacman::p_load(phyloregion, terra, sf, sp, tidyverse, ggplot2, ozmaps, tmap, viridis)

file_name <- str_split_1(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[2], "/")[10] |> str_remove(".parquet")

species_polys_points_expert <- readRDS(here("output/spatial", paste0("Spatial_alpha_hulls_sf_class_expert_", file_name, ".rds")))

# separating polygons and points for plotting
polys_ex <- species_polys_points_expert |>
  filter(st_geometry_type(conr_ahull_sf) %in% c("POLYGON", "MULTIPOLYGON"))
points_ex <- species_polys_points_expert |>
  filter(st_geometry_type(conr_ahull_sf) %in% c("POINT", "MULTIPOINT"))

# plotting alpha hulls and points
aus <- st_transform(ozmaps::ozmap_country, 4326)

## Base map 
base_map <- ggplot() + 
  geom_sf(data = aus, fill = "white", colour = "black") + 
  theme_minimal()

## alphahulls and points
ahulls_w_points <- base_map  + 
  geom_sf(data = polys_ex, 
          aes(geometry = conr_ahull_sf),
          fill = "grey",
          alpha = 0.5,
          colour = "darkgrey",
          ) +
  geom_sf(data = points_ex,
             aes(geometry = conr_ahull_sf),
             colour = "black",
             fill = "black") +
  geom_sf(data = aus, fill = NA, colour = "black") + 
  coord_sf(c(110, 155)) +
  theme_minimal() |> 
  theme(legend.position = "none")

```

```{r, echo=FALSE}
#| fig-cap: 
#|   - "Alpha-hulls of Mygalomorph spiders from our cleaned dataset, with point-based records (species with < 3 records) included as black points"

## alpha hulls
ahulls_w_points


# ## Community Matrix
# #|   #| fig-cap: 
# #|   - "Community matrix of Mygalomorphae spiders from our cleaned dataset"
# comm.poly <- phyloregion::polys2comm(dat = combinedShp , trace=1, res = 0.05)
# plot(comm.poly$map)
```

## Biodiversity metrics

::: {.callout-tip title="Source code"}
The biodiversity metrics workflow can be found [here](https://github.com/invertsau/inverthotspot/blob/main/R/05_endemism_mygalomorph.Rmd)
:::

The alphahull polygons and point-based records were transformed to Albers equal area projection (EPSG:9473) using the `project()` function from the `terra` package. In order to quantify patterns of species richness and endemism, these species distributuion data were converted into a community matrix, whereby each row in the matrix corresponds to a different site (spatial cell), while each column represents a different species. This was done using the `polys2comm()` function in the `phyloregion` R package. The entries in the matrix indicate whether a species is present in a particular cell, with a 1 for presence and 0 for absence. The resolution or grid size was set at 5 x 5 km. See [here](https://cran.r-project.org/web//packages/phyloregion/vignettes/phyloregion-intro.html) for further information.

There are multiple metrics to assess which areas are of importance for Mygalomorph spiders, each with their own caveats. In order to get a fuller picture of biodiversity and, importantly, rare species that are limited in their range, it is helpful to use a combination of metrics.

### Species richness

Species richness simply refers to the total number of species in an area. The `polys2comm()` function computes species richness for each cell.

### Endemism

Endemism refers to the level of range restriction of species in an area, and numerous measures exist (refs). Weighted endemism is species richness inversely weighted by species ranges (Crisp et al. 2001),(Laffan and Crisp 2003),(Barnabas H. Daru et al. 2020). It gives higher weight to species with smaller ranges, emphasizing areas with many range-restricted species. WE highlights areas with a high concentration of rare species, which can be crucial for conservation efforts. There is a caveat that regions with high species richness can have inflated WE values simply because they have more species, not necessarily more endemics (https://www.jstor.org/stable/3554498). However, WE is a good measure for identification and research into the distribution and conservation of endemism hotspots (Shipley and McGuire 2022). 

WE was calculated using the `weighted_endemism()` function from the `phyloregion` R package. WE was fourth-root transformed prior to mapping to aid visualization.

## Correcting for sampling bias

Sampling bias was estimated using a combination of sampling rate obtained from the `sampbias` package, and population density for the 2023-24 financial year (Australian Bureau of Statistics. (2023-24). Regional population. ABS. https://www.abs.gov.au/statistics/people/population/regional-population/2023-24.). Based on an input of species observation data, the `calculate_bias()` function from `sampbias` estimates sampling rate as a function of distance to geographic features including roads, cities, airports, and waterbodies. We included all of these features in the analysis and used a resolution of 0.05 decimal degrees. The sampling rate estimates were mapped using the `project_bias()` and `map_bias()` functions from `sampbias`. The sampling rate data were then extracted and log(+1) transformed to reduce skewness, before converting to a raster using the `rast()` function from the `terra` package. The community matrix created earlier was also converted to a raster.

The sampling rate estimates from `sampbias` showed the influence of regional centres such as Alice Springs being similar to large metropolitan areas such as Sydney. This is unlikely, so the effect of population density was also incorporated. Sampling rate and population density were transformed to Albers equal area projection (EPSG:9473) using the `project()` function from `terra`. Sampling rate and population density were then aligned to the community matrix raster using the `resample()` function from `terra` with bilinear interpolation. Population density was also log(+1) transformed to reduce skewness. The influence of population density was estimated using a gaussian smoothing kernel with a sigma value of 50 km (influence becomes near zero at 3*sigma), which was implemented via the `focal()` function in `terra`. The sampling rate from `sampbias` and population density effect were then each standardised by subtracting the mean and dividing by the standard deviation. The two measures were then combined into a single measure of sampling bias by taking the average. Finally, this total sampling bias measure was normalised to a range between 1-10. To correct for sampling bias, for each spatial cell, the biodiversity measures (richness and endemism) were divided by this final sampling bias estimate.

```{r, echo = FALSE}
#| fig-cap: 
#|   - "Maps of sampling bias estimates"

pop_density_influence <- readRDS(here("output/sampling_bias/pop_density_influence.rds"))
sampbias_sampling_rate <- readRDS(here("output/sampling_bias/sampbias_sampling_rate.rds"))
final_sampling_bias <- readRDS(here("output/sampling_bias/final_sampling_bias.rds"))

plot(c(sampbias_sampling_rate, pop_density_influence, final_sampling_bias))
```

## Hotspots

'Hotspots' were identified for each of the biodiversity metrics as the cells with values greater than the 95th percentile.

## Short-range species

The above analyses provide a quantification of endemism across all mygalomorph spiders, however, there is the potential for the influence of high-endemism species to be masked by the influence of widely distributed species. As short-range endemics are of particular conservation concern due to their higher susceptibility to threats such as habitat loss, the analyses were repeated including only species with a range size less than or equal to 10,000 km2 (i.e. occupying < 401 cells).

## Current protection

To assess the current level of protection of mygalomorph spider biodiversity, the overlap between hotspots and protected areas was calculated. Spatial polygons of protected areas in Australia were downloaded and cleaned using the `wdpar` R package. The proportion of species richness and endemism hotspots overlapping protected areas was then calculated and mapped. All maps were created using the `ggplot2` R package.

## Data and code availability

All data and code to reproduce the R portion of our analyses can be found at our [Github repository](https://github.com/invertsau/inverthotspot).
