---
execute:
  echo: false
  freeze: auto
editor_options: 
  chunk_output_type: console
---

# Results

```{r setup, include=FALSE, eval=TRUE}
## Setting up
 pacman::p_load(arrow, here, tidyverse, janitor, ozmaps, sf, ggplot2,sp, purrr, ConR, dplyr, DT, zoomerjoin, gt, patchwork, tmap, spdep, raster, terra)

tmap_mode("view")

# Load helper functions
source(here("R/utils.R"))

### Load in data
myg_spiders_cleaned <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[2])
myg_spiders_expert <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[1]) 
myg_spiders_hum_obs <- myg_spiders_cleaned |>
  filter(basisOfRecord == "HUMAN_OBSERVATION")

species_polys_points <- readRDS(here("output/spatial/Spatial_alpha_hulls_sf_class_cleaned_Mygalomorphae_withassertions_2025-08-14_ALA.rds"))
species_polys_points_expert <- readRDS(here("output/spatial/Spatial_alpha_hulls_sf_class_expert_cleaned_expert_recordsMygalomorphae_withassertions_2025-08-14_ALA.rds"))
species_polys_points_hum_obs <- readRDS(here("output/spatial/Spatial_alpha_hulls_sf_class_human_obs_cleaned_expert_recordsMygalomorphae_withassertions_2025-08-14_ALA.rds"))
```

## Preserved specimen-only data 

### All species

#### Hotspots

```{r, include = FALSE}
expert_hotspots <- read_csv(here("data/dataframes_for_plots/expert_hotspots.csv"))

# Expert richness
fig1a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = richness_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

# Expert endemism
fig1b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = WE_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")

# Expert richness hotspots
fig1c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = richness_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(expert_hotspots, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom")

# Expert endemism hotspots
fig1d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = WE_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(expert_hotspots, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") 
```

```{r}
#| fig-width: 12
#| fig-height: 12
#| fig-cap: 
#|   - "Figure 1: Maps showing species richness (left) and weighted endemism (right) of Mygalmorph spiders. The top row shows just the raw values and the bottom row has hotspots overlaid in red. Weighted endemism values were highly skewed so a fourth-root transformation was applied to aid visualisation."
fig1a + fig1b + fig1c + fig1d +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```

#### Current protection
```{r, include = FALSE}
aus_pa_data <- readRDS(here("data/protected_areas/aus_pa_data.rds"))

# clipping PA data to Aus coastline
wdpa_crs <- "+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs"

aus <- st_transform(ozmaps::ozmap_country, crs = st_crs(wdpa_crs))

aus_pa_data <-
  aus_pa_data %>%
  filter(MARINE == "terrestrial") %>%
  st_intersection(aus) 

# bringing in hotspot raster
merged_geo_ex <- rast(here("data/protected_areas/ex_bias_corrected_hotspots_spatraster.tif"))

# hotspots (95th percentiles)
rich_q95_ex <- global(merged_geo_ex[["richness_ex_sampbias_corrected"]], quantile, probs = 0.95, na.rm = TRUE)[1, 1]
merged_geo_ex$rich_perc <- merged_geo_ex[["richness_ex_sampbias_corrected"]] > rich_q95_ex

WE_q95_ex <- global(merged_geo_ex[["WE_ex_sampbias_corrected"]], quantile, probs = 0.95, na.rm = TRUE)[1, 1]
merged_geo_ex$WE_perc <- merged_geo_ex[["WE_ex_sampbias_corrected"]] > WE_q95_ex


v <- vect(aus_pa_data)
v <- project(v, merged_geo_ex)

aus_pa_raster <- rasterize(v, merged_geo_ex, touches = TRUE)

values(aus_pa_raster)[is.nan(values(aus_pa_raster))] <- NA


# getting cells where hotspots overlap with PA's
overlap_ex <- (!is.na(aus_pa_raster)) & (merged_geo_ex[["rich_perc"]] > 0)
names(overlap_ex) <- "richness"
overlap_ex$WE <- (!is.na(aus_pa_raster)) & (merged_geo_ex[["WE_perc"]] > 0)


# converting to data frame for plotting
aus_pa_raster_df <- as.data.frame(aus_pa_raster, xy = TRUE, na.rm = TRUE)
merged_geo_ex_df <- as.data.frame(merged_geo_ex, xy = TRUE, na.rm = TRUE)
overlap_ex_df <- as.data.frame(overlap_ex, xy = TRUE, na.rm = TRUE)


rich_hot_pa <- freq(overlap_ex$richness)[2,3]/freq(merged_geo_ex$rich_perc)[2,3]*100 # percentage of richness hotspots within PA's
we_hot_pa <- freq(overlap_ex$WE)[2,3]/freq(merged_geo_ex$WE_perc)[2,3]*100 # percentage of WE hotspots within PA's

# combining layers into one df for plotting
richness_tiles_df <- dplyr::bind_rows(
  dplyr::mutate(
    subset(merged_geo_ex_df, rich_perc),
    layer = "Hotspot"
  ),
  dplyr::mutate(
    aus_pa_raster_df,
    layer = "Protected area"
  ),
  dplyr::mutate(
    subset(overlap_ex_df, richness),
    layer = "Overlap"
  )
)

WE_tiles_df <- dplyr::bind_rows(
  dplyr::mutate(
    subset(merged_geo_ex_df, WE_perc),
    layer = "Hotspot"
  ),
  dplyr::mutate(
    aus_pa_raster_df,
    layer = "Protected area"
  ),
  dplyr::mutate(
    subset(overlap_ex_df, WE),
    layer = "Overlap"
  )
)

fig2a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = richness_tiles_df, aes(x=x, y=y, fill = layer), col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  scale_fill_manual(
    name = "",
    values = c(
      "Hotspot"     = "red",
      "Protected area" = "grey",
      "Overlap"        = "darkgreen"
    )) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

fig2b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = WE_tiles_df, aes(x=x, y=y, fill = layer), col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  scale_fill_manual(
    name = "",
    values = c(
      "Hotspot"     = "red",
      "Protected area" = "grey",
      "Overlap"        = "darkgreen"
    )) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: 
#|   - "Figure 2: Maps showing overlap between hotspots of mygalomorph spider species richness (left) and weighted endemism (right) with protected areas in Australia."
fig2a + fig2b +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```

::: callout-note
`r rich_hot_pa |> round(2)`% of the total species richness hotspot area and `r we_hot_pa |> round(2)`% of the total weighted endemism hotspot area are located within protected areas.
:::

### Short-range species

#### Hotspots

```{r, include = FALSE}
expert_hotspots_sr <- read_csv(here("data/dataframes_for_plots/expert_hotspots_sr.csv"))

# Expert richness
fig3a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = richness_sr_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

# Expert endemism
fig3b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = WE_sr_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")

# Expert richness hotspots
fig3c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = richness_sr_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(expert_hotspots_sr, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom")

# Expert endemism hotspots
fig3d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = WE_sr_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(expert_hotspots_sr, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") 
```

```{r}
#| fig-width: 12
#| fig-height: 12
#| fig-cap: 
#|   - "Figure 1: Maps showing species richness (left) and weighted endemism (right) of Mygalmorph spiders with short ranges (< 10,000 km2). The top row shows just the raw values and the bottom row has hotspots overlaid in red. Weighted endemism values were highly skewed so a fourth-root transformation was applied to aid visualisation."
fig3a + fig3b + fig3c + fig3d +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```

#### Current protection

```{r, include = FALSE}
aus_pa_data <- readRDS(here("data/protected_areas/aus_pa_data.rds"))

# clipping PA data to Aus coastline
wdpa_crs <- "+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs"

aus <- st_transform(ozmaps::ozmap_country, crs = st_crs(wdpa_crs))

aus_pa_data <-
  aus_pa_data %>%
  filter(MARINE == "terrestrial") %>%
  st_intersection(aus) 

# bringing in hotspot raster
merged_geo_sr_ex <- rast(here("data/protected_areas/sr_ex_bias_corrected_hotspots_spatraster.tif"))

# hotspots (95th percentiles)
rich_q95_sr_ex <- global(merged_geo_sr_ex[["richness_sr_ex_sampbias_corrected"]], quantile, probs = 0.95, na.rm = TRUE)[1, 1]
merged_geo_sr_ex$rich_perc <- merged_geo_sr_ex[["richness_sr_ex_sampbias_corrected"]] > rich_q95_sr_ex

WE_q95_sr_ex <- global(merged_geo_sr_ex[["WE_sr_ex_sampbias_corrected"]], quantile, probs = 0.95, na.rm = TRUE)[1, 1]
merged_geo_sr_ex$WE_perc <- merged_geo_sr_ex[["WE_sr_ex_sampbias_corrected"]] > WE_q95_sr_ex


v <- vect(aus_pa_data)
v <- project(v, merged_geo_sr_ex)

aus_pa_raster <- rasterize(v, merged_geo_sr_ex, touches = TRUE)

values(aus_pa_raster)[is.nan(values(aus_pa_raster))] <- NA


# getting cells where hotspots overlap with PA's
overlap_sr_ex <- (!is.na(aus_pa_raster)) & (merged_geo_sr_ex[["rich_perc"]] > 0)
names(overlap_sr_ex) <- "richness"
overlap_sr_ex$WE <- (!is.na(aus_pa_raster)) & (merged_geo_sr_ex[["WE_perc"]] > 0)


# converting to data frame for plotting
aus_pa_raster_df <- as.data.frame(aus_pa_raster, xy = TRUE, na.rm = TRUE)
merged_geo_sr_ex_df <- as.data.frame(merged_geo_sr_ex, xy = TRUE, na.rm = TRUE)
overlap_sr_ex_df <- as.data.frame(overlap_sr_ex, xy = TRUE, na.rm = TRUE)


sr_rich_hot_pa <- freq(overlap_sr_ex$richness)[2,3]/freq(merged_geo_sr_ex$rich_perc)[2,3]*100 # percentage of richness hotspots within PA's
sr_we_hot_pa <- freq(overlap_sr_ex$WE)[2,3]/freq(merged_geo_sr_ex$WE_perc)[2,3]*100 # percentage of WE hotspots within PA's


# combining layers into one df for plotting
richness_sr_tiles_df <- dplyr::bind_rows(
  dplyr::mutate(
    subset(merged_geo_sr_ex_df, rich_perc),
    layer = "Hotspot"
  ),
  dplyr::mutate(
    aus_pa_raster_df,
    layer = "Protected area"
  ),
  dplyr::mutate(
    subset(overlap_sr_ex_df, richness),
    layer = "Overlap"
  )
)

WE_sr_tiles_df <- dplyr::bind_rows(
  dplyr::mutate(
    subset(merged_geo_sr_ex_df, WE_perc),
    layer = "Hotspot"
  ),
  dplyr::mutate(
    aus_pa_raster_df,
    layer = "Protected area"
  ),
  dplyr::mutate(
    subset(overlap_sr_ex_df, WE),
    layer = "Overlap"
  )
)

fig4a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = richness_sr_tiles_df, aes(x=x, y=y, fill = layer), col = NA) +
   geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  scale_fill_manual(
    name = "",
    values = c(
      "Hotspot"     = "red",
      "Protected area" = "grey",
      "Overlap"        = "darkgreen"
    )) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

fig4b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = WE_sr_tiles_df, aes(x=x, y=y, fill = layer), col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  scale_fill_manual(
    name = "",
    values = c(
      "Hotspot"     = "red",
      "Protected area" = "grey",
      "Overlap"        = "darkgreen"
    )) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: 
#|   - "Figure 2: Maps showing overlap between hotspots of short-range (< 10,000 km2) mygalomorph spider species richness (left) and weighted endemism (right) with protected areas in Australia."
fig4a + fig4b +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```

::: callout-note
`r sr_rich_hot_pa |> round(2)`% of the species richness hotspot area and `r sr_we_hot_pa |> round(2)`% of the weighted endemism hotspot area for short-range species (< 10,000 km2) are located within protected areas.
:::

## Data source comparison

### All species

```{r, include = FALSE}
expert_hotspots <- read_csv(here("data/dataframes_for_plots/expert_hotspots.csv"))
human_obs_hotspots <- read_csv(here("data/dataframes_for_plots/hum_obs_hotspots.csv"))
full_dataset_hotspots <- read_csv(here("data/dataframes_for_plots/full_dataset_hotspots.csv"))

#defining a common richness scale for plotting
rich_limits <- range(
  expert_hotspots$richness_ex_sampbias_corrected,
  human_obs_hotspots$richness_hum_obs_sampbias_corrected,
  full_dataset_hotspots$richness_sampbias_corrected,
  na.rm = TRUE
)

richness_scale <-
  viridis::scale_fill_viridis(
    name = "Species richness (bias corrected)",
    option = "G",
    direction = -1,
    limits = rich_limits
  )

endemism_scale <-
  viridis::scale_fill_viridis(
    name = "Weighted endemism (bias corrected, fourth-root)",
    option = "G",
    direction = -1
  )

# Expert richness hotspots
fig5a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = richness_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(expert_hotspots, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

# Expert endemism hotspots
fig5b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots, aes(x=x, y=y, fill = WE_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(expert_hotspots, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")


# human observation richness hotspots
fig5c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = human_obs_hotspots, aes(x=x, y=y, fill = richness_hum_obs_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(human_obs_hotspots, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale +
  theme_minimal() +
  theme(legend.position = "bottom")

# human observation endemism hotspots
fig5d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = human_obs_hotspots, aes(x=x, y=y, fill = WE_hum_obs_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(human_obs_hotspots, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale +
  theme_minimal() +
  theme(legend.position = "bottom")


# full dataset richness hotspots
fig5e <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = full_dataset_hotspots, aes(x=x, y=y, fill = richness_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(full_dataset_hotspots, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale +
  theme_minimal() +
  theme(legend.position = "bottom")

# full dataset endemism hotspots
fig5f <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = full_dataset_hotspots, aes(x=x, y=y, fill = WE_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(full_dataset_hotspots, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| fig-width: 12
#| fig-height: 18
#| fig-cap: 
#|   - "Figure 5: Maps showing species richness (left) and weighted endemism (right) of Mygalmorph spiders based on different datasets: Expert data (basis of record == PRESERVED SPECIMEN) (top row); Citizen science (basis of record == HUMAN OBSERVATION) (middle row); Full dataset (bottom row). Hotspots are overlaid in red. Weighted endemism values were highly skewed so a fourth-root transformation was applied to aid visualisation."
fig5a + fig5b + fig5c + fig5d + fig5e + fig5f +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```

### Short-range species

```{r, include = FALSE}
expert_hotspots_sr <- read_csv(here("data/dataframes_for_plots/expert_hotspots_sr.csv"))
human_obs_hotspots_sr <- read_csv(here("data/dataframes_for_plots/hum_obs_hotspots_sr.csv"))
full_dataset_hotspots_sr <- read_csv(here("data/dataframes_for_plots/full_dataset_hotspots_sr.csv"))

#defining common scales for plotting
rich_limits_sr <- range(
  expert_hotspots_sr$richness_sr_ex_sampbias_corrected,
  human_obs_hotspots_sr$richness_sr_hum_obs_sampbias_corrected,
  full_dataset_hotspots_sr$richness_sr_sampbias_corrected,
  na.rm = TRUE
)

richness_scale_sr <-
  viridis::scale_fill_viridis(
    name = "Species richness (bias corrected)",
    option = "G",
    direction = -1,
    limits = rich_limits_sr
  )

WE_limits_sr <- range(
  expert_hotspots_sr$WE_sr_ex_sampbias_corrected,
  human_obs_hotspots_sr$WE_sr_hum_obs_sampbias_corrected,
  full_dataset_hotspots_sr$WE_sr_sampbias_corrected,
  na.rm = TRUE
)

endemism_scale_sr <-
  viridis::scale_fill_viridis(
    name = "Weighted endemism (bias corrected, fourth-root)",
    option = "G",
    direction = -1,
    limits = WE_limits_sr
  )

# Expert richness hotspots
fig6a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = richness_sr_ex_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(expert_hotspots_sr, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Species richness")

# Expert endemism hotspots
fig6b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = expert_hotspots_sr, aes(x=x, y=y, fill = WE_sr_ex_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(expert_hotspots_sr, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18)) +
  ggtitle("Weighted endemism")


# human observation richness hotspots
fig6c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = human_obs_hotspots_sr, aes(x=x, y=y, fill = richness_sr_hum_obs_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(human_obs_hotspots_sr, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom")

# human observation endemism hotspots
fig6d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = human_obs_hotspots_sr, aes(x=x, y=y, fill = WE_sr_hum_obs_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(human_obs_hotspots_sr, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom")


# full dataset richness hotspots
fig6e <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = full_dataset_hotspots_sr, aes(x=x, y=y, fill = richness_sr_sampbias_corrected), col = NA) + 
  labs(fill = "Species richness (bias corrected)") +
  geom_tile(data = subset(full_dataset_hotspots_sr, rich_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  richness_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom")

# full dataset endemism hotspots
fig6f <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_tile(data = full_dataset_hotspots_sr, aes(x=x, y=y, fill = WE_sr_sampbias_corrected^0.25), col = NA) + 
  labs(fill = "Weighted endemism (bias corrected)^0.25") +
  geom_tile(data = subset(full_dataset_hotspots_sr, WE_perc), aes(x=x, y=y), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155),
          ylim = c(-10, -45)) +
  endemism_scale_sr +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| fig-width: 12
#| fig-height: 18
#| fig-cap: 
#|   - "Figure 6: Maps showing species richness (left) and weighted endemism (right) of short-range (< 10,000 km2) Mygalmorph spiders based on different datasets: Expert data (basis of record == PRESERVED SPECIMEN) (top row); Citizen science (basis of record == HUMAN OBSERVATION) (middle row); Full dataset (bottom row). Hotspots are overlaid in red. Weighted endemism values were highly skewed so a fourth-root transformation was applied to aid visualisation."
fig6a + fig6b + fig6c + fig6d + fig6e + fig6f +
  plot_layout(ncol = 2,
              guides = 'collect') &
  theme(legend.position='bottom')
```