---
execute:
  echo: false
  freeze: auto
editor_options: 
  chunk_output_type: console
---

# Results

```{r setup, include=FALSE, eval=TRUE}
## Setting up
 pacman::p_load(arrow, here, tidyverse, janitor, ozmaps, sf, ggplot2,sp, purrr, ConR, dplyr, DT, zoomerjoin, gt, patchwork, tmap, spdep)

tmap_mode("view")

# Load helper functions
source(here("R/utils.R"))

### Load in data
myg_spiders_cleaned <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[2])
myg_spiders_expert <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[1]) 
myg_spiders_hum_obs <- myg_spiders_cleaned |>
  filter(basisOfRecord == "HUMAN_OBSERVATION")

species_polys_points <- readRDS("output/spatial/Spatial_alpha_hulls_sf_class_cleaned_Mygalomorphae_withassertions_2025-08-14_ALA.rds")
species_polys_points_expert <- readRDS("output/spatial/Spatial_alpha_hulls_sf_class_expert_cleaned_expert_recordsMygalomorphae_withassertions_2025-08-14_ALA.rds")
species_polys_points_hum_obs <- readRDS("output/spatial/Spatial_alpha_hulls_sf_class_human_obs_cleaned_expert_recordsMygalomorphae_withassertions_2025-08-14_ALA.rds")
```

## Alpha hulls

### Pre-processing

#### full dataset

```{r, include = FALSE}
# Extract POLYGONs from GEOMETRYCOLLECTIONs
species_polys_clean <- species_polys |>
  st_collection_extract() |>
  group_by(species) |>
  summarise(geometry = st_union(x))

combinedShp <- terra::vect(species_polys_clean)

### Calculate Community Matrix
comm.poly <- phyloregion::polys2comm(dat = combinedShp , trace=1, res = 0.05)
str(comm.poly)
plot(comm.poly$map)

### Calulate weighted endemism 
Endm.mygalomorph.poly <- phyloregion::weighted_endemism(comm.poly$comm_dat)
str(Endm.mygalomorph.poly)

## Join results back to spatial community data matrix
m1.poly <- merge(comm.poly$map, data.frame(grids=names(Endm.mygalomorph.poly), WE=Endm.mygalomorph.poly), by="grids")

# Drop NA in WE
m1.poly <- m1.poly[!is.na(m1.poly$WE),]

m1.poly

### Calulate corrected weighted endemism
m1.poly$corrected_endemism <- m1.poly$WE/m1.poly$richness

### Assigning hotspots based on percentile
m1.poly$rich_perc <- m1.poly$richness > quantile(m1.poly$richness, probs = 0.95)
m1.poly$WE_perc <- m1.poly$WE > quantile(m1.poly$WE, probs = 0.95)
m1.poly$CWE_perc <- m1.poly$corrected_endemism > quantile(m1.poly$corrected_endemism, probs = 0.95)

# Creating sf object
m1_sf <- 
  sf::st_as_sf(m1.poly) |> 
  sf::st_transform(4326)
```

#### expert-only data
```{r, include = FALSE}
# Extract POLYGONs from GEOMETRYCOLLECTIONs
species_polys_expert_clean <- species_polys_expert |>
  st_collection_extract() |>
  group_by(species) |>
  summarise(geometry = st_union(x))
  
combinedShp_ex <- terra::vect(species_polys_expert_clean)

### Calculate Community Matrix
comm.poly_ex <- phyloregion::polys2comm(dat = combinedShp_ex, species = "species", trace=1, res = 0.05)
str(comm.poly_ex)
plot(comm.poly_ex$map)

### Calulate weighted endemism 
Endm.mygalomorph.poly_ex <- phyloregion::weighted_endemism(comm.poly_ex$comm_dat)
str(Endm.mygalomorph.poly_ex)

## Join results back to spatial community data matrix
m1.poly_ex <- merge(comm.poly_ex$map, data.frame(grids=names(Endm.mygalomorph.poly_ex), WE=Endm.mygalomorph.poly_ex), by="grids")

# Drop NA in WE
m1.poly_ex <- m1.poly_ex[!is.na(m1.poly_ex$WE),]

### Calulate corrected weighted endemism
m1.poly_ex$corrected_endemism <- m1.poly_ex$WE/m1.poly_ex$richness

### Assigning hotspots based on percentile
m1.poly_ex$rich_perc <- m1.poly_ex$richness > quantile(m1.poly_ex$richness, probs = 0.95)
m1.poly_ex$WE_perc <- m1.poly_ex$WE > quantile(m1.poly_ex$WE, probs = 0.95)
m1.poly_ex$CWE_perc <- m1.poly_ex$corrected_endemism > quantile(m1.poly_ex$corrected_endemism, probs = 0.95)

# creating sf object
m1_sf_ex <- 
  sf::st_as_sf(m1.poly_ex) |> 
  sf::st_transform(4326)
```

### Species richness

```{r, include = FALSE}
# All data
fig1a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = richness), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")

# Expert Richness
fig1b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = richness), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 1: Map showing species richness of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"
fig1a + fig1b + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```

::: {.callout-note}
There was **no evidence** for spatial clustering in **species richness** in either the combined dataset or in the preserved specimen only dataset (Table 1).
:::

Table 1. Summary statistics for species richness in both datasets

```{r}
tibble(dataset = c("Citizen Science + Preserved Specimen",
                   "Preserved Specimen only"),
       `z-statistic` = c(R_I$statistic |> signif(2),
                     R_I_ex$statistic |> signif(2)),
       `p-value` = c(R_I$p.value  |> signif(2),
             R_I_ex$p.value  |> signif(2)),
       `n species` = c(myg_spiders_cleaned$scientificName |> unique() |> length(),
                       myg_spiders_cleaned_expert$scientificName |> unique() |> length()),
       `n observations` = c(nrow(myg_spiders_cleaned),
                            nrow(myg_spiders_cleaned_expert))) |> 
  clean_names(case = "sentence") |> 
  gt()
```

#### Hotspots

```{r}
# All data
fig1c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = richness), col = NA) + 
  geom_sf(data = subset(m1_sf, rich_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")

# Expert Richness
fig1d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = richness), col = NA) + 
  geom_sf(data = subset(m1_sf_ex, rich_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 2: Map showing hotspots of species richness of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"
fig1c + fig1d + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```

### Weighted endemism

```{r, include=FALSE}
# All data
fig2a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = WE^0.25), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")


# Expert data
fig2b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = WE^0.25), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 3: Map showing weighted endemism of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"

fig2a + fig2b + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```

::: {.callout-note}
There was **some evidence** for spatial clustering in **weighted endemism** in the preserved specimen only dataset (Table 2).
:::

```{r,include=FALSE}
WE_I

WE_I_ex
```

Table 2. Test statistic and p value for weighted endemsim

```{r}
tibble(dataset = c("Citizen Science + Preserved Specimen",
                   "Preserved Specimen only"),
       `z-statistic` = c(WE_I$statistic |> signif(2),
                     WE_I_ex$statistic |> signif(2)),
       `p-value` = c(WE_I$p.value  |> signif(2),
             WE_I_ex$p.value  |> signif(2)),
       `n species` = c(myg_spiders_cleaned$scientificName |> unique() |> length(),
                       myg_spiders_cleaned_expert$scientificName |> unique() |> length()),
       `n observations` = c(nrow(myg_spiders_cleaned),
                            nrow(myg_spiders_cleaned_expert))) |> 
  clean_names(case = "sentence") |> 
  gt()
```

#### Hotspots

```{r}
# All data
fig2c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = WE^0.25), col = NA) + 
  geom_sf(data = subset(m1_sf, WE_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")

# Expert data
fig2d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = WE^0.25), col = NA) + 
  geom_sf(data = subset(m1_sf_ex, WE_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1#, values = c(0,0.01,0.1,1)
                              ) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 4: Map showing hotspots of weighted endemism of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"
fig2c + fig2d + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```

### Corrected weighted endemism

```{r}
# All data
fig3a <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = corrected_endemism^0.25), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")

# "Expert Correct Endemism"
fig3b <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = corrected_endemism^0.25), col = NA) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 5: Map showing corrected weighted endemism estimated of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"

fig3a + fig3b + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```

```{r,include=FALSE}
CWE_I

CWE_I_ex
```

::: {.callout-note}
There was **some evidence** for spatial clustering in **corrected weighted endemism** in the preserved specimen only dataset (Table 3).
:::

Table 3. Test statistic and p value for corrected weighted endemsim

```{r}
tibble(dataset = c("Citizen Science + Preserved Specimen",
                   "Preserved Specimen only"),
       `z-statistic` = c(CWE_I$statistic |> signif(2),
                     CWE_I_ex$statistic |> signif(2)),
       `p-value` = c(CWE_I$p.value  |> signif(2),
             CWE_I_ex$p.value  |> signif(2)),
       `n species` = c(myg_spiders_cleaned$scientificName |> unique() |> length(),
                       myg_spiders_cleaned_expert$scientificName |> unique() |> length()),
       `n observations` = c(nrow(myg_spiders_cleaned),
                            nrow(myg_spiders_cleaned_expert))) |> 
  clean_names(case = "sentence") |> 
  gt()
```

#### Hotspots

```{r}
# All data
fig3c <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf, aes(fill = corrected_endemism^0.25), col = NA) + 
  geom_sf(data = subset(m1_sf, CWE_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Citizen science + Preserved specimen data")

# "Expert Correct Endemism"
fig3d <- ggplot() +
  geom_sf(data = ozmap_states, fill = "white") + 
  geom_sf(data = m1_sf_ex, aes(fill = corrected_endemism^0.25), col = NA) + 
  geom_sf(data = subset(m1_sf_ex, CWE_perc), fill = "red", col = NA) +
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  ggtitle("Preserved specimen data only")
```

```{r}
#| fig-width: 8
#| fig-cap: 
#|   - "Figure 6: Map showing hotspots of corrected weighted endemism of Mygalmorphae spiders. Left: All data from the cleaned dataset is used i.e contains citizen science data (basis of record == `HUMAN_OBSERVATION`) Right: Only expert data is used (basis of record == `PRESERVED_SPECIMEN` only)"
fig3c + fig3d + 
  plot_layout(guides = 'collect') &
  theme(legend.position='bottom')
```