---
title: "myg assertions overview"
author: "Josh Nitschke"
date: "2023-11-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load dependencies

```{r loadpackages}
# install.packages("pacman")
pacman::p_load(galah, arrow, here, tidyverse, janitor, ozmaps, sf, ggplot2, wesanderson, dplyr)

```

## Read in parquet

```{r}
myg_spiders_all <- open_dataset(here("data/galah/Mygalomorphae_withassertions_2023-12-01_ALA.parquet"))
```

## Subset columns we want for ALA data overview

```{r}
myg_spiders <- myg_spiders_all |> 
  select(decimalLatitude:collectionCode) |> 
  collect()
```

### Overview of assertions

```{r}
myg_assertions <- myg_spiders_all |> 
  dplyr::select(scientificName, taxonRank, species, decimalLatitude:decimalLongitude, recordID, collectionCode, institutionCode, AMBIGUOUS_COLLECTION:ZERO_COORDINATE) |> 
  collect()

myg_assertions

myg_assertions_long <- myg_assertions |>
  pivot_longer(cols = c(AMBIGUOUS_COLLECTION:ZERO_COORDINATE), names_to = "assertion_type", values_to = "value") 

assertion_counts <- myg_assertions_long |>
  filter(value == TRUE) |>
  tabyl(assertion_type)

# Join with ala_assertions.csv so we have summary of counts
#asserts <- read_csv("output/ala_assertions.csv")
#asserts |> 
 # left_join(assertion_counts, by = join_by(id == assertion_type)) |>
 # arrange(-n) |> 
 # write_csv("output/ala_assertions_with_counts.csv")

#Filtering observations where all assertions are false
myg_assertions |>
  filter(!if_any(AMBIGUOUS_COLLECTION:ZERO_COORDINATE))
```

## read in assertion counts

```{r}
asserts <- read_csv("data/ala_assertions_with_counts.csv")

asserts |>
  filter(to_investigate == 1) |>
  arrange(type...8) |>
  print(n = Inf)
  
```

### Checking coordinate-related assertions
```{r}
myg_assertions |>
  filter(!is.na(species),
         !is.na(decimalLatitude),
         !is.na(decimalLongitude)) |>
  filter(if_any(starts_with("COORDINATE"))) |>
  group_by(species) |>
  tabyl(species) |>
  arrange(desc(n))

# plotting spp. with the most flagged records for coordinate-based assertions
myg_spp_coordinate_assertions <- myg_assertions |>
  filter(!is.na(species),
         !is.na(decimalLatitude),
         !is.na(decimalLongitude)) |>
  filter(if_any(starts_with("COORDINATE"))) |>
  group_by(species) |>
  filter(n() > 100)

aus <- st_transform(ozmaps::ozmap_states, 4326)

ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = myg_spp_coordinate_assertions, 
             mapping = aes(decimalLongitude, decimalLatitude, colour = species), 
             size = 0.8) + 
  # coord_sf(xlim=c(135, 155), 
  #          ylim=c(-10, -32)) +
  labs(title = "myg_spiders: assertions starting with 'COORDINATE' = TRUE", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw()

```

## Investigating spatial assertions

Coordinate uncertainty meters invalid
Keeping NAs and anything = or less than 1000 m uncertainty

spot check 10 records of the 46 flagged for 5 m
```{r}
coord_uncertainty_m_invalid <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, COORDINATE_UNCERTAINTY_METERS_INVALID)

coord_uncertainty_m_invalid_spp <- coord_uncertainty_m_invalid |>
  group_by(species) |>
  summarise(num_true = sum(COORDINATE_UNCERTAINTY_METERS_INVALID)) |>
  filter(num_true > 0) |>
  pull(species)

coord_uncertainty_m_invalid_spp_no_na <- myg_spiders_all |>
  filter(species %in% coord_uncertainty_m_invalid_spp,
         !is.na(species)) |>
  collect()

coord_uncertainty_m_invalid_spp_no_na |>
  filter(COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE)|>
  select(species, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters)|>
  tabyl(coordinateUncertaintyInMeters)

coord_uncertainty_m_invalid_spp_no_na |>
  filter(coordinateUncertaintyInMeters == 5) |>
  tabyl(species, COORDINATE_UNCERTAINTY_METERS_INVALID) 

coord_uncertainty_m_invalid_spp_no_na |>
  filter(species == "Atrax robustus",
         coordinateUncertaintyInMeters == 5) |>
  ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(mapping = aes(decimalLongitude, decimalLatitude, colour = COORDINATE_UNCERTAINTY_METERS_INVALID), 
             size = 1.5) + 
  #scale_colour_manual(values = wes_palette("BottleRocket2")) +
   coord_sf(xlim=c(150, 152), 
            ylim=c(-32, -35)) +
  labs(title = "Atrax robustus: coord uncertainty m = 5", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw()
  #facet_wrap(~assertion_type)

coord_uncertainty_m_invalid_spp_no_na |>
  filter(species == "Atrax robustus",
         coordinateUncertaintyInMeters == 5,
         COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE) |>
  select(decimalLatitude:dataResourceName, COORDINATE_UNCERTAINTY_METERS_INVALID, 
         coordinateUncertaintyInMeters, coordinatePrecision,
         basisOfRecord:collectionCode, recordID) |> #tabyl(coordinatePrecision)
  pull(recordID)

coord_uncertainty_m_invalid_spp_no_na |>
  filter(species == "Hadronyche versuta",
         coordinateUncertaintyInMeters == 5,
         COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE) |>
  select(decimalLatitude:dataResourceName, COORDINATE_UNCERTAINTY_METERS_INVALID, 
         coordinateUncertaintyInMeters, coordinatePrecision,
         basisOfRecord:collectionCode, recordID) |> #tabyl(coordinatePrecision)
  pull(recordID)

coord_uncertainty_m_invalid_spp_no_na |>
  filter(COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE)|>
  select(species, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, UNCERTAINTY_IN_PRECISION)|>
  tabyl(coordinateUncertaintyInMeters, UNCERTAINTY_IN_PRECISION)

myg_spiders_all |> 
  filter(!is.na(species)) |>
  #filter(COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE)|>
  select(species, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, UNCERTAINTY_IN_PRECISION)|>
  collect() |>
  tabyl(UNCERTAINTY_IN_PRECISION)

# seems that COORDINATE_UNCERTAINTY_METERS_INVALID == TRUE when UNCERTAINTY_IN_PRECISION == TRUE, and I think 
# UNCERTAINTY_IN_PRECISION == TRUE when coordinateUncertaintyInMeters is inferred from coordinatePrecision...
```

Coordinate rounded
```{r}
coord_rounded <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, COORDINATE_ROUNDED)

coord_rounded |>
  group_by(species) |>
  tabyl(species, COORDINATE_ROUNDED) |>
  arrange(desc(eval(as.name("TRUE"))))

coord_rounded_spp <- coord_rounded |>
  group_by(species) |>
  summarise(num_true = sum(COORDINATE_ROUNDED)) |>
  filter(num_true > 0) |>
  pull(species)

coord_rounded_spp_no_na <- myg_spiders_all |>
  filter(species %in% coord_rounded_spp,
         !is.na(species)) |>
  collect()

coord_rounded_spp_no_na |>
  filter(COORDINATE_ROUNDED == TRUE)|>
  select(species, decimalLatitude, decimalLongitude, recordID)|>
  view()
```


Coordinate invalid
```{r}
coord_invalid <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, COORDINATE_INVALID)

coord_invalid_spp <- coord_invalid |>
  group_by(species) |>
  summarise(num_true = sum(COORDINATE_INVALID)) |>
  filter(num_true > 0) |>
  pull(species)

coord_invalid_spp_no_na <- myg_assertions |>
  filter(species %in% coord_invalid_spp,
         !is.na(species))

  ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = coord_invalid_spp_no_na, mapping = aes(decimalLongitude, decimalLatitude, colour = COORDINATE_INVALID), 
             size = 0.8, alpha = 0.5) + 
  # coord_sf(xlim=c(135, 155), 
  #          ylim=c(-10, -32)) +
  labs(title = "myg_spiders: coordinate_invalid", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
    facet_wrap(~species)
  
coord_invalid_spp_no_na |>
  filter(COORDINATE_INVALID == TRUE)
```

get verbatum coordinates
```{r}
miss_occ_coord_flagged <- myg_assertions |>
  filter(species == "Missulena occatoria") |>
  select(species, decimalLatitude, decimalLongitude, contains("COORDINATE")) |>
  pivot_longer(cols = contains("COORDINATE"), names_to = "assertion_type", values_to = "assertion_flag") |>
  filter(assertion_flag == TRUE) 

ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = miss_occ_coord_flagged, mapping = aes(decimalLongitude, decimalLatitude, colour = assertion_type), 
             size = 1.5) + 
  scale_colour_manual(values = wes_palette("BottleRocket2")) +
  # coord_sf(xlim=c(135, 155), 
  #          ylim=c(-10, -32)) +
  labs(title = "myg_spiders: miss_occ_coord_flagged", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
  facet_wrap(~assertion_type)

```

### For rounded coordinates, investigate the raw ones

```{r}
# Configure display settings
old <- options(pillar.sigfig = 4)

myg_spiders_coordinate_rounded <- myg_spiders_all |> 
  select(contains("decimal"), contains("coord"), ends_with("Name"), recordID,
         starts_with("raw"), starts_with("verbatim"), COORDINATE_ROUNDED, COORDINATE_INVALID, UNCERTAINTY_IN_PRECISION, COORDINATE_UNCERTAINTY_METERS_INVALID) |> 
  collect()

myg_spiders_coordinate_rounded |> 
  filter(COORDINATE_ROUNDED == TRUE) |> 
  select(verbatimCoordinates, raw_decimalLatitude, raw_decimalLongitude, decimalLatitude, decimalLongitude) |> 
  print(n = 20)

rounde_myg <- myg_spiders_coordinate_rounded |> 
  filter(COORDINATE_ROUNDED == TRUE) 

rounde_myg$decimalLatitude 

# Trying to find those pesky rounded values
library(stringr)

# Match cases following decimal point with 000 after.
str_subset(as.character(rounde_myg$decimalLatitude), regex("(?<=\\.\\d{2})000")) 

examples <- rounde_myg |> 
  filter(recordID %in% c("6e0df496-ce6a-430c-a068-e68fe8c7f1e9",
                       "5f5cc21a-bace-425c-82cb-97d346b6d353",
                       "432e290b-67c5-4b0c-b5bb-9549c047be8d")) 

examples$decimalLatitude
examples$raw_decimalLatitude

# Early conclusion, some species are just truncated to 1 dp, there is also the setting you can adjust to display more dps/sfs

# Counts the number of digits in decimalLatitude
str_count(as.character(rounde_myg$decimalLatitude), "\\d") |> 
  tabyl()

# Add counts to tibble
rounde_myg |> 
  mutate(digit_counts = str_count(as.character(rounde_myg$decimalLatitude), "\\d")) |> 
  filter(digit_counts %in% c(2,3)) |> 
  select(decimalLatitude, raw_decimalLatitude, digit_counts, COORDINATE_ROUNDED, UNCERTAINTY_IN_PRECISION) |> 
  View()

# Conclusion: Can't replace rounded values with the raw_ coords, as they were already rounded.
```

Spatial assertions that are safe:

- COORDINATE_UNCERTAINTY_METERS_INVALID 
- COORDINATE_ROUNDED
- COORDINATE_OUT_OF_RANGE

Spatial assertions that we can use to exclude data:
- PRESUMED_NEGATED_LONGITUDE
- PRESUMED_SWAPPED_COORDINATE
- ZERO_COORDINATE

JM to investigate
- COORDINATES_CENTRE_OF_STATEPROVINCE (Likely to be excluded)
- COUNTRY_COORDINATE_MISMATCH (Likely to be excluded)
- STATE_COORDINATE_MISMATCH (Are these in borderlines, have they then been corrected by ALA, is there conflict between stateProvince an decimLat and decimLong, try redownloading with raw_stateProvince) 


COORDINATES_CENTRE_OF_STATEPROVINCE
```{r}
coord_centre_state <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, COORDINATES_CENTRE_OF_STATEPROVINCE)

coord_centre_state |>
  filter(COORDINATES_CENTRE_OF_STATEPROVINCE == TRUE) |>
  tabyl(species)

coord_centre_state_spp <- coord_centre_state |>
  group_by(species) |>
  summarise(num_true = sum(COORDINATES_CENTRE_OF_STATEPROVINCE)) |>
  filter(num_true > 0) |>
  pull(species)

coord_centre_state_spp_no_na <- myg_assertions |>
  filter(species %in% coord_centre_state_spp,
         !is.na(species))

ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = coord_centre_state_spp_no_na, mapping = aes(decimalLongitude, decimalLatitude, colour = COORDINATES_CENTRE_OF_STATEPROVINCE), 
             size = 1.5, alpha = 0.5) + 
   coord_sf(xlim=c(110, 160), 
            ylim=c(-10, -45)) +
  labs(title = "myg_spiders: coord_centre_state", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
    facet_wrap(~species)
  
# Only two species (one record each) and within the range of other records for those species, so probably not an issue

```

COUNTRY_COORDINATE_MISMATCH
```{r}
country_coord_mismatch <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, COUNTRY_COORDINATE_MISMATCH)

country_coord_mismatch |>
  filter(COUNTRY_COORDINATE_MISMATCH == TRUE) |>
  tabyl(species)

country_coord_mismatch_spp <- country_coord_mismatch |>
  group_by(species) |>
  summarise(num_true = sum(COUNTRY_COORDINATE_MISMATCH)) |>
  filter(num_true > 0) |>
  pull(species)

country_coord_mismatch_spp_no_na <- myg_spiders_all |>
  filter(species %in% country_coord_mismatch_spp,
         !is.na(species)) |>
  collect()

  ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = country_coord_mismatch_spp_no_na, 
              mapping = aes(decimalLongitude, decimalLatitude, colour = COUNTRY_COORDINATE_MISMATCH), 
              size = 1.5, alpha = 0.5) + 
   #coord_sf(xlim=c(110, 160), 
   #         ylim=c(-10, -45)) +
  labs(title = "myg_spiders: country_coord_mismatch", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
    facet_wrap(~species)

country_coord_mismatch_spp_no_na |> 
  select(species, decimalLatitude, decimalLongitude, country, COUNTRY_COORDINATE_MISMATCH)
```

Masking Australia mainland and Tas
```{r}
ozmaps::ozmap_country

myg_spiders_clean_sf <- myg_spiders |>
 select(species, decimalLatitude, decimalLongitude, country) |>
  filter(!is.na(decimalLatitude),
         !is.na(decimalLongitude),
         !is.na(species)) |>
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), # Set as sf object
                                      crs = 4326) |> 
                             rename(points = geometry)

ozmap_country_reprojected <- st_transform(ozmaps::ozmap_country, 4326)

x <- st_intersection(ozmap_country_reprojected, myg_spiders_clean_sf)

plot(x)
```


STATE_COORDINATE_MISMATCH
```{r}
state_coord_mismatch <- myg_assertions |>
  select(species, decimalLatitude, decimalLongitude, STATE_COORDINATE_MISMATCH)

state_coord_mismatch |>
  filter(STATE_COORDINATE_MISMATCH == TRUE) |>
  tabyl(species) |>
  arrange(desc(n))

state_coord_mismatch_spp <- state_coord_mismatch |>
  group_by(species) |>
  summarise(num_true = sum(STATE_COORDINATE_MISMATCH)) |>
  filter(num_true > 0) |>
  pull(species)

state_coord_mismatch_spp_no_na <- myg_spiders_all |>
  filter(species %in% state_coord_mismatch_spp,
         !is.na(species)) |>
  collect()

ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(data = state_coord_mismatch_spp_no_na, 
              mapping = aes(decimalLongitude, decimalLatitude, colour = STATE_COORDINATE_MISMATCH), 
              size = 1.5, alpha = 0.5) + 
   #coord_sf(xlim=c(110, 160), 
   #         ylim=c(-10, -45)) +
  labs(title = "myg_spiders: state_coord_mismatch", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
    facet_wrap(~species)

# filtering most flagged species
state_coord_mismatch_spp_no_na |>
  filter(species == "Bymainiella terraereginae") |>
  ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(mapping = aes(decimalLongitude, decimalLatitude, colour = STATE_COORDINATE_MISMATCH), 
              size = 1.5, alpha = 0.5) + 
   coord_sf(xlim=c(150, 155), 
            ylim=c(-27, -30)) +
  labs(title = "Bymainiella terraereginae: state_coord_mismatch", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw()

# filtering species with at least 5 records where STATE_COORDINATE_MISMATCH == TRUE
state_coord_mismatch_spp_five_or_more_true <- state_coord_mismatch |>
  group_by(species) |>
  summarise(num_true = sum(STATE_COORDINATE_MISMATCH)) |>
  filter(num_true > 4) |>
  pull(species)

state_coord_mismatch_spp_no_na |>
  filter(species %in% state_coord_mismatch_spp_five_or_more_true,
         STATE_COORDINATE_MISMATCH == TRUE) |>
  ggplot() + 
  geom_sf(data = aus, 
          colour = "black", 
          fill = "white")  + 
   geom_point(mapping = aes(decimalLongitude, decimalLatitude, colour = stateProvince), 
              size = 1.5, alpha = 0.5) + 
   coord_sf(xlim=c(110, 160), 
            ylim=c(-10, -45)) +
  labs(title = "myg_spiders: state_coord_mismatch", 
       x = "Longtitude (DD)", 
       y = "Latitude (DD)") +
  theme_bw() +
  facet_wrap(~species)

# TRUEs are mixed in with FALSEs and stateProvince seems to align with coordinates, so probably not an issue. Maybe the issue is with raw_stateProvince
```

Taxon match fuzzy: original and matched names next to each other
printing other taxon assertions (species and rawScientific name) in own chunk

TAXON_MATCH_FUZZY
```{r}
# Number of raw_scientificName variations for each species with fuzzy taxon matches
myg_spiders_all |>
  filter(TAXON_MATCH_FUZZY == TRUE) |>
  select(species, raw_scientificName, recordID) |>
  collect() |>
  group_by(species) |>
  summarise(num_raw_names = length(unique(raw_scientificName)))

# List of species (adjusted name) and raw names (names given in observation)
fuzzy_taxa <- myg_spiders_all |>
  filter(TAXON_MATCH_FUZZY == TRUE) |>
  select(species, scientificName, raw_scientificName) |>
  collect() |>
  unique() |>
  print(n = Inf)
```

TAXON_EXCLUDED
```{r}
# List of species (adjusted name) and raw names (names given in observation)
taxa_excluded <- myg_spiders_all |>
  filter(TAXON_EXCLUDED == TRUE) |>
  select(species, scientificName, raw_scientificName) |>
  collect() |>
  unique() |>
  print(n = Inf)
# seems that species has been made NA and scientificName reduced to genus

# counts of TAXON_EXCLUDED == TRUE 
myg_assertions |>
  filter(TAXON_EXCLUDED == TRUE) |>
  tabyl(scientificName)
```

TAXON_QUESTION_SPECIES
```{r}
# List of species (adjusted name) and raw names (names given in observation)
taxa_question <- myg_spiders_all |>
  filter(TAXON_QUESTION_SPECIES == TRUE) |>
  select(species, scientificName, raw_scientificName) |>
  collect() |>
  unique() |>
  print(n = Inf)

# counts of TAXON_QUESTION_SPECIES == TRUE 
myg_assertions |>
  filter(TAXON_QUESTION_SPECIES == TRUE) |>
  tabyl(scientificName)
```

TAXON_SPECIES_PLURAL
```{r}
# List of species (adjusted name) and raw names (names given in observation)
taxa_question <- myg_spiders_all |>
  filter(TAXON_SPECIES_PLURAL == TRUE) |>
  select(species, scientificName, raw_scientificName) |>
  collect() |>
  unique() |>
  print(n = Inf)

# counts of TAXON_SPECIES_PLURAL == TRUE 
myg_assertions |>
  filter(TAXON_SPECIES_PLURAL == TRUE) |>
  tabyl(scientificName)
```

TAXON_HOMONYM
```{r}
# List of species (adjusted name) and raw names (names given in observation)
taxa_question <- myg_spiders_all |>
  filter(TAXON_HOMONYM == TRUE) |>
  select(species, scientificName, raw_scientificName, recordID) |>
  collect() |>
  unique() |>
  print(n = Inf)
# these records are from Spain?...

# counts of TAXON_HOMONYM == TRUE 
myg_assertions |>
  filter(TAXON_HOMONYM == TRUE) |>
  tabyl(scientificName)
```