---
title: "Data cleaning"
author: "Ashley Browse"
date: "2023-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Cleaning data from ALA

This document lays out the workflow of cleaning data for Mygalomorph spiders from the ALA database 

## Load dependencies

```{r loadpackages}
# install.packages("pacman")
pacman::p_load(galah, arrow, here, tidyverse, janitor, dplyr, galah, worrms, ozmaps, sf, CoordinateCleaner)

```


## Read in parquet (observations retrieved from ALA)

```{r}
myg_spiders_all <- open_dataset(here("data/galah/Mygalomorphae_withassertions_2023-12-01_ALA.parquet"))
```


## Subset columns we want

```{r}
myg_spiders <- myg_spiders_all |> 
  select(decimalLatitude:outlierLayer) |> 
  collect()
```


### Taxonomic errors

<!-- Clean for taxonomic errors: misspelt names, incorrect, synonyms.  -->

### Filter down to species level data

Are species and scientificName not the same thing?? 

`species` and `scientificName` are the same for Mygalomorphae download

```{r}
myg_spiders |> 
  count(taxonRank) 

myg_species <- myg_spiders |> 
  filter(taxonRank == "species")

# Checking with new variable
myg_species |> 
  select(species, scientificName)  |> 
  mutate(exact_match = species == scientificName) |> 
  count(exact_match)
```


### Read in Australian Faunal Directory (AFD) data to cross check out taxa

The [Australian Faunal Directory (AFD)](https://biodiversity.org.au/afd/home) is an online catalogue of taxonomic and biological information on all animal species known to occur within Australia and its territories.

The AFD can provide information on:

- nomenclature and taxonomy of species (including valid names [= what a species is currently called], synonyms, and changes to names over time);
- type data information;
- bibliographic information for species;
- distribution of species.

<!-- Check names against a naming authority to make sure all species included in the analyses are valid -->

```{r}
# afd <- read_csv(here("data/afd_May2023_clean.csv"))
# afd |> write_parquet(here("data/afd_05-2023_clean.parquet"))
afd <- open_dataset(here("data/afd_05-2023_clean.parquet"))

# Filter down to Arachnida 
arachnids <- afd |> 
  filter(CLASS == "ARACHNIDA") |> 
  collect()
```

Where is Mygalomorphae in the AFD?

There in no infraorder column in the AFD, can filter down to Mygalomorphae using `HIGHER_CLASSIFICATION`
 
```{r}
arachnids |> 
  pull(HIGHER_CLASSIFICATION) |> 
  str_subset(pattern = "MYGALOMORPH") |> head()

names(arachnids)
```

How many matches in ALA data, do we have in the AFD?

```{r}
#x <- c(1, 2, 3, 6, 7, 8, 8, 9)
#y <- c(1, 2, 5, 4, 9)

#setdiff(x, y)
#x %in% y

# 488 species in ALA data matched in the AFD
myg_species |> 
  filter(scientificName %in% arachnids$FULL_NAME) |> 
  pull(scientificName) |> 
  unique() |> 
  length()

# How many and which ones are the species in ALA data that DID NOT match with the AFD
myg_species |> 
  filter(! scientificName %in% arachnids$FULL_NAME) |> 
  pull(scientificName) |> 
  unique() |> 
  length()

# Why are these species not in the AFD? These might be synonyms! 
unmatched_ala_sp <- myg_species |> 
  filter(! scientificName %in% arachnids$FULL_NAME) |> 
  pull(scientificName) |> 
  unique() 
unmatched_ala_sp
```


Which of the ALA species not matched to AFD occur in the list of AFD arachnid species synonyms

```{r}

# Pull out the species name from Synonyms without the Authority
# "Austrammo monteithi Platnick, 2002"     -> "Austrammo monteithi"
# word(), str_split()

# Using similar methods above try cross match unmatched_ala_sp to synonyms

arachnids$SYNONYMS |> head()
#remove Authority, keep species name
word(arachnids$SYNONYMS, 1, 2) |> head()

#check if unmatched species occur in list of synonyms
issynonym <- unmatched_ala_sp %in% word(arachnids$SYNONYMS, 1, 2)
issynonym

#One species occurs in synonyms list. Which one?
unmatched_ala_sp[issynonym]

```


<!-- Excluding introduced/invasives taxa -->

### Use World Spider Catalog and GRIIS to exclude introduced spiders from dataset

```{r}
# Read in spider data
wsc <- read_csv(here("data/wsc/species_export_20230929.csv"))
```

Have a look at distribution variable
```{r}
# `distribution` values that contain "Australia"
str_subset(wsc$distribution, "Australia")

# Return unique values of `distribution` for those that have "Introduced to Australia" in distribution
wsc |> 
  pull(distribution) |> str_subset(pattern = "Introduced to Australia") |>  
  unique()

wsc |> 
  pull(distribution) |> str_subset(pattern = "Australia") |>  
  unique()

# We want to capture all observations where "Introduced to" and "Australia" occurs together
str_subset(wsc$distribution, pattern = "Introduced to.*Australia")
str_detect(wsc$distribution, pattern = "Introduced to.*Australia")
str_which(wsc$distribution, pattern = "Introduced to.*Australia")

introduced_taxa <- wsc |> 
  filter(str_detect(distribution, pattern = "Introduced to.*Australia")) 

# We need to join genus and species together manually
introduced_taxa_wfullname <- introduced_taxa |> 
  mutate(scientific_name = paste(genus, species)) 

introduced_taxa_wfullname |> 
  select(scientific_name)
```

Cross match if introduced taxa from WSC with ALA data to exclude introduced spiders

There are no introduced species of spiders in ALA data according to WSC data

```{r}
introduced_taxa_wfullname$scientific_name %in% myg_species$scientificName |> table()
myg_species$scientificName %in% introduced_taxa_wfullname$scientific_name |> table()
```


### Global Register of Invasive and Introduced Species (GRIIS)

```{r}
# Join distrubtion.txt and taxon-edit.txt
griis_distrib <- read.delim(here("data/griis-australia-v1.6/distribution.txt"))
griis_taxon <- read.delim(here("data/griis-australia-v1.6/taxon-edited.txt"))
summary(griis_distrib) #length = 2984
summary(griis_taxon) #length = 2960
#which IDs are missed? (What is in Distribution that is not in Taxon?)
setdiff(griis_distrib$id, griis_taxon$id)
#is there anything missing from Distribution?
setdiff(griis_taxon$id, griis_distrib$id) # no there isn't

griis_all <- left_join(griis_distrib, griis_taxon, by = "id")                 # Apply left_join dplyr function 

#filter by class = arachnida
str_subset(griis_all$class, "Arachnida")
griis_all$class |> unique() #there are no arachnids in this list


# Cross match with ALA data


```



<!-- Excluding marines -->

<!-- Notes from Payal -->
<!-- Decide whether to keep morpho-species  -->
<!-- How do you distinguish these in ALA data? -->
<!-- FK to check with PB, I don't think these in are ALA data -->
```{r}

```


### Subspecies

<!-- Decide whether to keep sub-species and if not whether to combine their data with the parent species -->
 
```{r}
#look for subspecies
myg_spiders$taxonRank |> unique() #what is "unranked" and "NA"
myg_spiders |> filter(taxonRank == "unranked") #no species IDs
myg_spiders |> filter(taxonRank == "NA") #nothing
#look at unique species
myg_spiders$species |> unique() #no subspecies here



```


### Excluding Marine Species

```{r, eval=FALSE}
wm_records_taxamatch("Teranodes")
Myg_species <- myg_spiders$species |> unique()




job::job({
  output_df <- map(Myg_species,
                   possibly(~wm_records_taxamatch(.x, marine = FALSE)  |>
                              pluck(1)  |>
                              mutate(search_term = .x) |>
                              discard(.p = ~is.null(.x))
                   )
  ) |>
    list_rbind()

  saveRDS(output_df, "output/worrms_myg_wm_records_taxamatch.rds")
})
```


### Geographic errors

<!-- Clean for geographic errors: missing or invalid lat/long; if data needs to be masked by land/water. There are also R packages like coordinate cleaner and others (you'd know more) to clean occurrence data  -->

Remove NA is decimalLat and decimalLong

```{r}
myg_species_aus <- myg_species |> 
  filter(decimalLatitude != "NA",
         decimalLongitude != "NA") #remove NA coordinates

summary(myg_species_aus$decimalLatitude)
summary(myg_species_aus$decimalLongitude) #NAs gone
```

Restricting data to Australia mainland + Tasmania

```{r}
myg_species_mainland <- myg_species_aus |> 
  filter(decimalLatitude > -45, 
         decimalLatitude < -8, 
         decimalLongitude > 112, 
         decimalLongitude < 154) 

# Transform into sf object
myg_species_mainland_sf <- myg_species_mainland |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326)


# Verified that we have cropped the data
# Map data
aus <- st_transform(ozmaps::ozmap_country, 4326)

base_map <- ggplot() + 
  geom_sf(data = aus, fill = NA) + 
  theme_minimal()

base_map  + 
  geom_sf(data = myg_species_mainland_sf) + 
  theme_minimal()
```

What are the species we've excluded from the mainland? 

```{r}

```

Coordinate Cleaner

https://docs.ropensci.org/CoordinateCleaner/

`.equ` = Equal longitude and latitude
`.sea` = XXX 

```{r}
flagged <- clean_coordinates(x = myg_species_mainland, 
                                     "decimalLongitude",
                                     "decimalLatitude")

# Summary of how many flagged records found for each category
summary(flagged)

plot(flagged, lon = "decimalLongitude", lat = "decimalLatitude")
```

### Visualise .sea values

FALSE are values that are flagged to occur in sea and therefore will be eventually excluded


```{r}
flagged_sf <- flagged |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326) 

base_map  + 
  geom_sf(data = flagged_sf, aes(color = .sea), alpha = 0.3) + 
  theme_minimal()
```

### Visualise .cap values

### Visualise .otl values

### Visualise .inst values




<!-- Check for duplicates (as the last step) -->

In this step we can inspect and clean the data of duplicates 

```{r}
nrow(myg_spiders)
head(myg_spiders)

Mygalomorphae_clean <- myg_spiders |> 
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) |>
  filter(!duplicated(decimalLatitude) & !duplicated(decimalLongitude)) 

nrow(Mygalomorphae_clean)
head(Mygalomorphae_clean)
```

