---
title: "Alpha hulls"
author: "Fonti Kar and Payal Bal"
date: "2023-08-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadpackages}
# Install the package
#install.packages("remotes")
# remotes::install_github("beatrizpateiro/alphahull", ref="WIP-ahull.IUCN-feature")
# remotes::install_github("babichmorrowc/hull2spatial")
# remotes::install_github("gdauby/ConR")

# Load packages
#install.packages("pacman")
pacman::p_load(galah, arrow, here, tidyverse, ozmaps, sf, ggplot2, skimr, alphahull, hull2spatial, sp, purrr, ConR, dplyr)
```

## Read in data

```{r}
myg_spiders <- read_parquet(here("data/galah/Mygalomorphae_withassertions_2023-09-18_ALA.parquet"))
```

## Precleaning

A quick clean for now, data cleaning is still being finalised

```{r}
myg_spiders_cleaned <- myg_spiders |> 
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) |>
  filter(!duplicated(decimalLatitude) & !duplicated(decimalLongitude)) 
```

## Narrow down to species with enough observations

```{r}
# Filter down to species
myg_cleaned_sp <- myg_spiders_cleaned |> 
  filter(taxonRank == "species")

# Summarise number of observations by taxon
myg_cleaned_sp |> 
  group_by(scientificName) |> 
  summarise(n = n()) |> 
  arrange(-n)

# Which ones have less than or equal 3 
less_than_3_obs <- myg_cleaned_sp |> 
  group_by(scientificName) |> 
  summarise(n = n()) |> 
  filter(n <= 3) |> 
  pull(scientificName)

myg_sp_enoughobs <- myg_cleaned_sp |> 
  filter(!scientificName %in% less_than_3_obs)

# How many species
myg_sp_enoughobs |>
  dplyr::select(scientificName) |> 
  distinct() |> 
  nrow()
```

## Proof of concept

Selecting a random taxa to create alphahulls, so we can abstract the workflow to all taxa

```{r}
set.seed(987)
choosen <- myg_sp_enoughobs |> 
  slice_sample(n = 1) |> 
  pull(scientificName)

b_aurea <- myg_sp_enoughobs |> 
  filter(scientificName == choosen)

b_aurea |> skim()
```

### A quick overview of observations

Make a simple map

```{r}
# Transform projection
aus <- st_transform(ozmaps::ozmap_country, 4326)

# Base map 
base_map <- ggplot() + 
  geom_sf(data = aus, fill = "white") + 
  theme_minimal()

# Transform spiders into sf objects
b_aurea_sf <- b_aurea |> 
  select(scientificName, decimalLongitude, decimalLatitude) |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326)

# Base map 
base_map  + 
  geom_sf(data = b_aurea_sf, alpha = 0.5) + 
  theme_minimal()
```

## Create an alpha hull

Alpha hulls are a flexible way to represent species distributions without all the extra work of running a species distribution model. 

There are many ways to create an alphahull, this way is Bea's method which was borrowed by the `ConR` R package which was used in the original NESP project.

After sharing the [IUCN documentation, page 51](https://nc.iucnredlist.org/redlist/content/attachment_files/RedListGuidelines.pdf) with Bea, she noted that her method is different what is stated in the documentation [(Burgman and Fox 2003)](https://espace.library.uq.edu.au/view/UQ:39112/Bias_species_ranges_2003.pdf). 

After discussing the potential impact of the IUCN alpha hull for the conservation/biodiversity community, Bea agreed to implement a function to estimate alpha hulls using the Burgman and Fox method. [This work](https://github.com/beatrizpateiro/alphahull/tree/WIP-ahull.IUCN-feature) is a work in progress and not yet released to public. 

## Bea's version from {alphahull} package

```{r}
# Create alphaull
b_aurea_ahull <- b_aurea |> 
  select(decimalLongitude, decimalLatitude) |> 
  ahull(alpha = 2)

plot(b_aurea_ahull)

# Transform to sf object
b_aurea_ahull_sf <- ahull2poly(b_aurea_ahull) |>
  st_as_sf() |>
  st_set_crs(4326)  # This method is not working not sure why

# Another way to transform to sf object
# b_aurea_ahull_2 <- hull2spatial:::ahull2lines(b_aurea_ahull) |> st_as_sf() |> st_cast(to = "MULTIPOLYGON")  |> st_set_crs(4326)

# Add to our map
base_map  + 
  geom_sf(data = b_aurea_sf, alpha = 0.5) + 
  geom_sf(data = b_aurea_ahull_sf, fill = 'orange', alpha = 0.2) +
  theme_minimal()
```
### Extroplate across all taxa

```{r}
create_ahull_sf <- function(data, 
                            lat, long, 
                            alpha = 2){
  
# browser()
ahull_obj <- data |> 
  dplyr::select({{long}},  {{lat}}) |> 
  alphahull::ahull(alpha = alpha)

# Convert to sp then sf
ahull_sf <- hull2spatial::ahull2poly(ahull_obj) |>
  sf::st_as_sf() |>
  sf::st_set_crs(4326)

return(ahull_sf)

}
# Test this out
create_ahull_sf(b_aurea, decimalLatitude, decimalLongitude)
```

### Nest data and try apply function across all taxa

```{r}
myg_nested_sp <- myg_sp_enoughobs |> 
  select(scientificName, decimalLongitude, decimalLatitude) |> 
  nest(occ = c(decimalLongitude, decimalLatitude))

myg_nested_ahull <- myg_nested_sp |> 
  mutate(points_sf = map(.x = occ,
                           ~ st_as_sf(.x, coords = c("decimalLongitude", "decimalLatitude"), # Set as sf object
                                      crs = 4326) |> 
                             rename(points = geometry)), # Rename geometry variable to something intuitive
         ahulls_sf = map(myg_nested_sp$occ, 
    possibly(~create_ahull_sf(.x, decimalLatitude, decimalLongitude))
  )
  )

# This throws an error
myg_unnested <- myg_nested_ahull |> 
  select(scientificName, points_sf, ahulls_sf) |> 
  unnest(cols = c(points_sf, ahulls_sf)) 

# Bind manually
ahull_bound <- list_rbind(myg_nested_ahull$ahulls_sf) |> 
  rename(ahull = geometry)

# Identify which ones are null
null_hulls_taxa <- which(lengths(myg_nested_ahull$ahulls_sf) == 0)

# Investigate NULL taxa 
# Not sure why these are NULL for ahulls, some dups and some very similar values
myg_nested_ahull |> 
  slice(null_hulls_taxa) |> 
  pluck("occ")

problem_taxa <- myg_nested_ahull |> 
  slice(null_hulls_taxa) |> 
  pull(scientificName)

# Left join in other taxonomic info
with_taxonomic_info <- myg_nested_ahull |> left_join(myg_sp_enoughobs |> 
  select(scientificName, genus, family)) |> 
  distinct()

# Exclude these for now
myg_hull_plot <- with_taxonomic_info |> 
  filter(!scientificName %in% problem_taxa) |> 
  select(-occ) |>
  unnest(c(points_sf, ahulls_sf)) 


# Cropping data to mainland Australia
myg_hull_plot |> 
   # filter(decimalLatitude < -10, decimalLatitude >= -45,
   #       decimalLongitude >= 113, decimalLongitude <= 155) |> 
  st_as_sf(crs = 4326) 
```

### Plot the ahulls

```{r}
base_map + # There are offshore points need to crop these
    geom_sf(data = myg_hull_plot, aes(geometry = points), colour = "black", 
              alpha = 0.1)

base_map + # There are offshore points need to crop these
    geom_sf(data = myg_hull_plot, aes(geometry = points), colour = "black", 
              alpha = 0.2) + 
  geom_sf(data = myg_hull_plot, aes(geometry = geometry), fill = "#023E50",
          alpha = 0.01) +
  coord_sf(ylim=c(-10,-45),
         xlim=c(113,155)) 

base_map + # There are offshore points need to crop these
    geom_sf(data = myg_hull_plot, aes(geometry = points), colour = "black", 
              alpha = 0.2) + 
  geom_sf(data = myg_hull_plot, aes(geometry = geometry, fill = family),
          alpha = 0.01) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.5))) + 
  coord_sf(ylim=c(-10,-45),
         xlim=c(113,155)) + 
  theme(legend.position = "bottom")
```

## IUCN version

We need to install alphahull from the [IUCN branch](https://github.com/beatrizpateiro/alphahull/tree/WIP-ahull.IUCN-feature) to use these in code

```{r}
# remotes::install_github("beatrizpateiro/alphahull", ref = "WIP-ahull.IUCN-feature")
library(alphahull)
```

```{r}
# Create alphaull
b_aurea_ahull_iucn <- b_aurea |> 
  select(decimalLongitude, decimalLatitude) |> 
  alphahull::ahull.IUCN(alpha = 4)

plot(b_aurea_ahull_iucn) # Yikes this looks funky hey!

# Complete the structure between methods
str(b_aurea_ahull_iucn)
str(b_aurea_ahull)


# Transform to sf object
ahull2poly(b_aurea_ahull_iucn) # Yeah I didn't think it would be that easy
```

### Update {hull2spatial} code

```{r}
# hull2spatial:::ahull2lines # Prints the function
# Expected behaviour
hull2spatial:::ahull2lines(b_aurea_ahull)

hull = b_aurea_ahull
hull_2 = b_aurea_ahull_iucn

IA_ahull2lines <- function(hull) 
{
  # Determine what type of ahull it is
  if(class(hull) ==  "ahull.IUCN")
    arclist <- hull$arcs
  
  arclist <- hull$arcs
  lines <- list()
  for (i in 1:nrow(arclist)) {
    center_i <- arclist[i, 1:2]
    radius_i <- arclist[i, 3]
    vector_i <- arclist[i, 4:5]
    theta_i <- arclist[i, 6]
    line_i <- arc2line(center = center_i, r = radius_i, vector = vector_i, 
                       theta = theta_i)
    list_length <- length(lines)
    if (list_length > 0) {
      last_line_coords <- lines[[list_length]]@coords
    }
    if (i == 1) {
      lines[[i]] <- line_i
    }
    else if (isTRUE(all.equal(line_i@coords[1, ], last_line_coords[nrow(last_line_coords), 
    ]))) {
      lines[[list_length]]@coords <- rbind(last_line_coords, 
                                           line_i@coords[2:nrow(line_i@coords), ])
    }
    else {
      lines[[length(lines) + 1]] <- line_i
    }
  }
  lines <- Lines(lines, ID = "l")
  sp_lines <- SpatialLines(list(lines))
  return(sp_lines)
}

```

### ConR

#### Proof of concept for one sp 

```{r}
b_aurea 

# Reshuffle data according to documentation requirements
b_aurea_conr <- b_aurea |> 
  dplyr::select(decimalLatitude, decimalLongitude, scientificName)

conr_output <- IUCN.eval(b_aurea_conr, 
                         method.range = "alpha.hull",
                         alpha = 2, 
                         export_shp = TRUE, 
                         write_shp = TRUE)

conr_output 

# Read in shp
b_aurea_shp <- st_read(here("shapesIUCN/Blakistonia_aurea_EOO_poly.shp")) |> 
  st_set_crs(value = 4326)

plot(b_aurea_shp)
```

#### Extrapolate to all Myg taxa

```{r}
myg_conr_nest <- myg_sp_enoughobs |> 
  dplyr::select(decimalLatitude, decimalLongitude, scientificName) |> 
  mutate(species = scientificName) |> 
  nest(occ = c(starts_with("decimal"), scientificName))

# A function to map over
map(.x = myg_conr_nest$occ, 
    possibly(~IUCN.eval(DATA =.x, 
                        method.range = "alpha.hull",
                        alpha = 2, 
                        export_shp = TRUE, 
                        write_shp = TRUE)
    )
)


# With Buffer
myg_concr_output_withbuffer <- map(.x = myg_conr_nest$occ, 
    possibly(~IUCN.eval(DATA =.x, 
                        method.range = "alpha.hull",
                        alpha = 2, 
                        export_shp = TRUE, 
                        )
    )
)

write_rds(myg_concr_output_withbuffer, here("output/myg_concr_output_withbuffer_0.1.rds"))

# No buffer
myg_concr_output <- map(.x = myg_conr_nest$occ, 
    possibly(~IUCN.eval(DATA =.x, 
                        method.range = "alpha.hull",
                        alpha = 2, 
                        export_shp = TRUE, 
                        buff.alpha = 	0.000001 # This is in decimal degrees why ~	111 mm
                        )
    )
)

# write_rds(myg_concr_output, here("output/myg_concr_output.rds"))

myg_concr_output <- read_rds(here("output/myg_concr_output.rds"))

# Flatten output
myg_concr_output <- myg_concr_output |> flatten() 

# Need a mechanism to name the list
myg_concr_output[[1]]$Results |> names()

length(myg_concr_output) == nrow(myg_conr_nest)

names(myg_concr_output) <- myg_conr_nest$species

myg_concr_output

# Extract spatialPoly_EOO only for endemism workflow
get_EOO_sp <- function(sp_name){
  output <- myg_concr_output |> 
  pluck(sp_name, "spatialPoly_EOO") 
}

myg_spiders_conr_sp <- map(.x = myg_conr_nest$species,
     ~get_EOO_sp(.x))

length(myg_spiders_conr_sp) == nrow(myg_conr_nest)

names(myg_spiders_conr_sp) <- myg_conr_nest$species

write_rds(myg_spiders_conr_sp, "output/myg_spiders_conr_sp.rds")


# A function to pull spatialPoly_EOO, change it to sf and add in species name
get_ahull_sf <- function(sp_name){
  output <- myg_concr_output |> 
  pluck(sp_name, "spatialPoly_EOO")  |> 
  st_as_sf() |> 
  mutate(species = sp_name) |> 
  dplyr::select(-dummy) 
  
  row.names(output) <- NULL
  
  output
}

myg_spiders_conr_sf <- map_df(.x = myg_conr_nest$species,
     ~get_ahull_sf(.x))

write_rds(myg_spiders_conr_sf, "output/myg_sf_alphahulls.rds")
```

### Plot these all at once

```{r}
myg_spiders_conr_sf <- readRDS("output/myg_sf_alphahulls.rds")

base_map + # There are offshore points need to crop these
  geom_sf(data = myg_spiders_conr_sf, aes(geometry = geometry), fill = "#023E50",
          alpha = 0.1) +
  coord_sf(ylim=c(-10,-45),
         xlim=c(113,155)) + 
  theme_minimal()

base_map + # There are offshore points need to crop these
    geom_sf(data = myg_hull_plot, aes(geometry = points), colour = "black", 
              alpha = 0.2) + 
  geom_sf(data = myg_hull_plot, aes(geometry = geometry, fill = family),
          alpha = 0.01) +
  guides(fill = guide_legend(override.aes = list(alpha = 0.5))) + 
  coord_sf(ylim=c(-10,-45),
         xlim=c(113,155)) + 
  theme(legend.position = "bottom")
```

A few example taxa

```{r}
test <- myg_spiders_conr_sf |>  
  sample_frac(0.2) 

test
test_occ <- myg_conr_nest |> 
  filter(species %in% test$species) |> 
  unnest() |> 
  dplyr::select(-scientificName) |> 
  st_as_sf(crs = 4326, coords = c("decimalLongitude", "decimalLatitude")) 


base_map + # There are offshore points need to crop these
  geom_sf(data = test, aes(geometry = geometry), fill = "#023E50",
          alpha = 0.2) +
  geom_sf(data = test_occ, size = 0.2, colour = "deepskyblue2") +
  coord_sf(ylim=c(-10,-45),
         xlim=c(113,155)) + 
  theme_minimal()
```
 
devtools::install_github("gdauby/ConR")

myg_spiders <- read_parquet(here("data/galah/Mygalomorphae_withassertions_2023-09-18_ALA.parquet"))



dataset <- dummy_dist(nsp = 1, max_occ = 30)

alphas <- EOO.computing(XY = dataset, method.range = "alpha", export_shp = TRUE, alpha = 3)

plot(alphas$spatial)

alphas <- EOO.computing(XY = dataset, method.range = "alpha", export_shp = TRUE, alpha = 3, buff.alpha = 	0.000001)

plot(alphas$spatial)


