---
title: "Alpha hulls"
author: "Fonti Kar and Payal Bal"
date: "2023-08-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadpackages}
# Install the package
#install.packages("remotes")
# remotes::install_github("beatrizpateiro/alphahull", ref="WIP-ahull.IUCN-feature")
# remotes::install_github("babichmorrowc/hull2spatial")

# Load packages
#install.packages("pacman")
pacman::p_load(galah, arrow, here, tidyverse, ozmaps, sf, ggplot2, skimr, alphahull, hull2spatial, sp, purrr, ConR, dplyr)

source("R/utils.R")
```

## Read in data

```{r}
get_latest_cleaned_data()

myg_spiders_cleaned <- read_parquet(get_latest_cleaned_data()) 
```

## Proof of concept

Selecting a random taxa to create alphahulls, so we can abstract the workflow to all taxa

```{r}
set.seed(987)
choosen <- myg_spiders_cleaned |> 
  slice_sample(n = 1) |> 
  pull(scientificName)

b_aurea <- myg_spiders_cleaned |> 
  filter(scientificName == choosen)

b_aurea |> skim()
```

### A quick overview of observations

Make a simple map

```{r}
# Transform projection
aus <- st_transform(ozmaps::ozmap_country, 4326)

# Base map 
base_map <- ggplot() + 
  geom_sf(data = aus, fill = "white") + 
  theme_minimal()

# Transform spiders into sf objects
b_aurea_sf <- b_aurea |> 
  select(scientificName, decimalLongitude, decimalLatitude) |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326)

# Base map 
base_map  + 
  geom_sf(data = b_aurea_sf, alpha = 0.5) + 
  theme_minimal()
```

## Create an alpha hull

Alpha hulls are a flexible way to represent species distributions without all the extra work of running a species distribution model. 

### ConR

#### Proof of concept for one sp 

```{r}
b_aurea 

# Reshuffle data according to documentation requirements
b_aurea_conr <- b_aurea |> 
  dplyr::select(decimalLatitude, decimalLongitude, scientificName)

conr_output <- EOO.computing(b_aurea_conr, 
                             method.range = "alpha",
                             alpha = 2, 
                             buff.alpha = 0.000001,
                             export_shp = TRUE,
                             write_shp = TRUE)

as(conr_output$spatial, "Spatial")
```

#### Extrapolate to all Myg taxa

```{r}
# A function to generate alpha-hull and extract output and transform into sf object
ConR_ahull <- function(data, sf_transform = TRUE){

  conr_output <- ConR::EOO.computing(data, 
                               method.range = "alpha",
                               alpha = 2, 
                               buff.alpha = 0.000001,
                               export_shp = TRUE)
  if(sf_transform == TRUE){
  as(conr_output$spatial, "Spatial")
  } 
  else
    conr_output$spatial
}

# Nest data
myg_conr_nest <- myg_spiders_cleaned |> 
  dplyr::select(decimalLatitude, decimalLongitude, scientificName, eventDate) |> 
  mutate(species = scientificName) |> 
  nest(occ = c(starts_with("decimal"), scientificName, eventDate))

# Test function on one
test <- ConR_ahull(myg_conr_nest$occ[[1]], sf_transform = TRUE)


# A function to map over
#Can't get this map working for some reason!
myg_conr_nest <- myg_conr_nest |>
  mutate(conr_ahull = purrr::map(.x = occ,
                          possibly(~ConR_ahull(.x))
  ),
  conr_ahull_sf =  purrr::map(.x = occ,
                          possibly(~ConR_ahull(.x, FALSE))
  )) 
```

### Drop empty polygons

```{r}
myg_conr_nest |> 
  print(n = Inf)

x <- myg_conr_nest |> 
  purrr::map(discard, .p = is.null) |> 
  tibble()

null_spdf <- myg_conr_nest$conr_ahull |> 
  purrr::map(is.null) |> 
  list_c() |> which()

myg_conr_nest |> 
  # filter(row_number() %in% null_spdf) |> 
  pluck("occ", 2)
```

### Alpha hulls as sf to plot all on a map

```{r}
# Transform projection
aus <- st_transform(ozmaps::ozmap_country, 4326)

# Base map 
base_map <- ggplot() + 
  geom_sf(data = aus, fill = NA, colour = "black") + 
  theme_minimal()

#Unnest the sf
all_myg_alpha_sf <- myg_conr_nest |> 
  select(species, conr_ahull_sf) |> 
  tidyr::unnest(cols = c(conr_ahull_sf)) |> 
  select(-tax) |> 
  sf::st_as_sf()

# Save output
# Clean up the name
file_name <- str_split_1(get_latest_cleaned_data(), "/")[3] |> str_remove(".parquet")

saveRDS(all_myg_alpha_sf, here("output/spatial", paste0("Spatial_alpha_hulls_sf_class_", file_name, ".rds")))

# Base map 
base_map  + 
  geom_sf(data = all_myg_alpha_sf, 
          aes(geometry = x),
           # colour = "black",
          colour = "#609966") + 
  coord_sf(c(110, 155)) +
  theme_minimal() |> 
  theme(legend.position = "none")

```

#### Save output as rds for next step of workflow

```{r}
get_latest_cleaned_data()

# Clean up the name
file_name <- str_split_1(get_latest_cleaned_data(), "/")[3] |> str_remove(".parquet")

# Save spatial dataframe
write_rds(myg_conr_nest$conr_ahull, here("output/spatial", paste0("Spatial_alpha_hulls_", file_name, ".rds")))

myg_concr_output <- read_rds(here("output/myg_concr_output.rds"))
```



