---
title: "Endemism analysis"
author: "Fonti Kar, Payal Bal, Aaron Greenfield"
date: "2023-12-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(phyloregion, terra, sf, sp, tidyverse, ggplot2, ozmaps, tmap, viridis, spdep, purrr)
```

### Load alpha hull spatial polygons

This is a massive list, each element is a spatial polygon dataframe

```{r}
species_polys <- readRDS("output/spatial/filled_alpha_hulls_sf_cleaned_Mygalomorphae_withassertions_2024-03-13_ALA.rds")

# species_polys_expert <- readRDS("output/spatial/Spatial_alpha_hulls_expertcleaned_Mygalomorphae_withassertions_2024-03-13_ALA.rds") |> 
#   discard(is.null)
```

## ALL DATA

### Pre-processing

```{r}

# Combine into one large spatial dataframe
combinedShp <- bind_rows(species_polys, .id = "species")

combinedShp <- terra::vect(combinedShp)

combinedShp

```

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` which generates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces" the function; trace = 2 or higher will be more voluminous

```{r}
comm.poly <- phyloregion::polys2comm(dat = combinedShp , trace=1, res = 0.5)
str(comm.poly)
plot(comm.poly$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.poly <- phyloregion::weighted_endemism(comm.poly$comm_dat)
str(Endm.mygalomorph.poly)

## Join results back to spatial community data matrix
m1.poly <- merge(comm.poly$map, data.frame(grids=names(Endm.mygalomorph.poly), WE=Endm.mygalomorph.poly), by="grids")

# Drop NA in WE
m1.poly <- m1.poly[!is.na(m1.poly$WE),]

m1.poly
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.poly$corrected_endemism <- m1.poly$WE/m1.poly$richness
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf <- 
  sf::st_as_sf(m1.poly) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf, aes(fill = WE)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf, aes(fill = corrected_endemism)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf, aes(fill = richness)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#:~:text=Moran's%20I%20values%20significantly%20above,negative%20spatial%20autocorrelation%20or%20dispersion.

```{r}
# extract the center of each polygon
coo <- terra::centroids(m1.poly)

coo <- coo |> sf::st_as_sf(crs = 4326)

# Search radius to include all neighboring polygon (0 - 200km)
S.dist  <-  dnearneigh(coo, 0, 200000)  

#identify all neighboring polygons for each polygon in the dataset.
lw <- nb2listw(S.dist, style="W",zero.policy=T) 

# Run the MC simulation
RI  <-  moran.mc(m1.poly$richness, lw, nsim=9999,zero.policy=T)
MI  <-  moran.mc(m1.poly$corrected_endemism, lw, nsim=9999,zero.policy=T)
WE_MI  <-  moran.mc(m1.poly$WE, lw, nsim=9999,zero.policy=T)

saveRDS(RI, "output/analysis/Cleaned_Mygalomorphae_RI.rds")
saveRDS(MI, "output/analysis/Cleaned_Mygalomorphae_MI.rds")
saveRDS(WE_MI, "output/analysis/Cleaned_Mygalomorphae_WE_MI.rds")
```

## EXPERT DATA ONLY


```{r}
# Combine into one large spatial dataframe
combinedShp_ex <- do.call(what = rbind, args=species_polys_expert)

## Assign species names
combinedShp_ex <- terra::vect(combinedShp_ex)
names(combinedShp_ex) <- "species"

combinedShp_ex
```

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` which generates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces" the function; trace = 2 or higher will be more voluminous

```{r}
comm.poly_ex <- phyloregion::polys2comm(dat = combinedShp_ex, species = "species", trace=1, res = 0.5)
str(comm.poly_ex)
plot(comm.poly_ex$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.poly_ex <- phyloregion::weighted_endemism(comm.poly_ex$comm_dat)
str(Endm.mygalomorph.poly_ex)

## Join results back to spatial community data matrix
m1.poly_ex <- merge(comm.poly_ex$map, data.frame(grids=names(Endm.mygalomorph.poly_ex), WE=Endm.mygalomorph.poly_ex), by="grids")

# Drop NA in WE
m1.poly_ex <- m1.poly_ex[!is.na(m1.poly_ex$WE),]

m1.poly_ex
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.poly_ex$corrected_endemism <- m1.poly_ex$WE/m1.poly_ex$richness
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf_ex <- 
  sf::st_as_sf(m1.poly_ex) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = WE)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = corrected_endemism)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = richness)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#:~:text=Moran's%20I%20values%20significantly%20above,negative%20spatial%20autocorrelation%20or%20dispersion.

```{r}
# extract the center of each polygon
coo <- terra::centroids(m1.poly_ex)

coo <- coo |> sf::st_as_sf(crs = 4326)

# Search radius to include all neighboring polygon (0 - 200km)
S.dist  <-  dnearneigh(coo, 0, 200000)  

#identify all neighboring polygons for each polygon in the dataset.
lw <- nb2listw(S.dist, style="W",zero.policy=T) 

# Run the MC simulation
RI_ex  <-  moran.mc(m1.poly_ex$richness, lw, nsim=9999,zero.policy=T)
MI_ex <-  moran.mc(m1.poly_ex$corrected_endemism, lw, nsim=9999,zero.policy=T)
WE_MI_ex  <-  moran.mc(m1.poly_ex$WE, lw, nsim=9999,zero.policy=T)

saveRDS(RI_ex, "output/analysis/Cleaned_Mygalomorphae_RI_expert.rds")
saveRDS(MI_ex, "output/analysis/Cleaned_Mygalomorphae_MI_expert.rds")
saveRDS(WE_MI_ex, "output/analysis/Cleaned_Mygalomorphae_WE_MI_expert.rds")
```
