---
title: "Endemism analysis"
author: "Fonti Kar, Payal Bal, Aaron Greenfield"
date: "2023-12-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(galah, arrow, here, phyloregion, terra, sf, sp, tidyverse, ggplot2, ozmaps, tmap, viridis, spdep, purrr)

source("R/utils.R")
```

## Read in data

```{r}
get_latest_cleaned_data()

myg_spiders_cleaned <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[2])

myg_spiders_expert <- read_parquet(get_latest_cleaned_data(path_to_clean_data = paste0(here("output/data/"), "/"))[1]) 
```

## ALL DATA

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` which generates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces" the function; trace = 2 or higher will be more voluminous

```{r}
comm.point <- phyloregion::points2comm(dat = myg_spiders_cleaned, res = 0.5, lon = "decimalLongitude", lat = "decimalLatitude", species = "species")
str(comm.point)
plot(comm.point$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.point <- phyloregion::weighted_endemism(comm.point$comm_dat)
str(Endm.mygalomorph.point)

## Join results back to spatial community data matrix
m1.point <- merge(comm.point$map, data.frame(grids=names(Endm.mygalomorph.point), WE=Endm.mygalomorph.point), by="grids")

# Drop NA in WE
m1.point <- m1.point[!is.na(m1.point$WE),]

m1.point
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.point$corrected_endemism <- m1.point$WE/m1.point$richness
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf <- 
  sf::st_as_sf(m1.point) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf, aes(fill = WE)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf, aes(fill = corrected_endemism)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf, aes(fill = richness)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#:~:text=Moran's%20I%20values%20significantly%20above,negative%20spatial%20autocorrelation%20or%20dispersion.

```{r}
# extract the center of each polygon
coo <- terra::centroids(m1.point)

coo <- coo |> sf::st_as_sf(crs = 4326)

st_is_longlat(coo)

# Search radius to include all neighboring polygon (0 - 200km)
S.dist  <-  dnearneigh(coo, 0, 200)  

#identify all neighboring polygons for each polygon in the dataset.
lw <- nb2listw(S.dist, style="W",zero.policy=T) 

# Run the MC simulation
R_MI_point  <-  moran.mc(m1.point$richness, lw, nsim=9999,zero.policy=T)
WE_MI_point  <-  moran.mc(m1.point$WE, lw, nsim=9999,zero.policy=T)
CWE_MI_point  <-  moran.mc(m1.point$corrected_endemism, lw, nsim=9999,zero.policy=T)

# saveRDS(R_MI_point, "output/analysis/Cleaned_Mygalomorphae_point_R_MI.rds")
# saveRDS(WE_MI_point, "output/analysis/Cleaned_Mygalomorphae_point_WE_MI.rds")
# saveRDS(CWE_MI_point, "output/analysis/Cleaned_Mygalomorphae_point_CWE_MI.rds")

```

Local Moran's I - to identify significant clusters
```{r}
tmap_mode("view")

RI_local <- localmoran(m1.point$richness, lw, alternative = "greater")

m1_sf$lmI <- RI_local[, "Ii"] # local Moran's I
m1_sf$lmZ <- RI_local[, "Z.Ii"] # z-scores
m1_sf$lmp <- RI_local[, "Pr(z > E(Ii))"] # p-values corresponding to alternative

# moran's I scatterplot
mp <- moran.plot(as.vector(scale(m1_sf$richness)), lw)

#designating points to quadrants based on their position in the moran's scatterplot and the significance of local moran's I
m1_sf$quadrant <- NA
# high-high
m1_sf[(mp$x >= 0 & mp$wx >= 0) & (!is.na(m1_sf$lmp) & m1_sf$lmp <= 0.0001), "quadrant"]<- 1
# low-low
m1_sf[(mp$x <= 0 & mp$wx <= 0) & (!is.na(m1_sf$lmp) & m1_sf$lmp <= 0.0001), "quadrant"]<- 2
# non-significant
m1_sf[(!is.na(m1_sf$lmp) & m1_sf$lmp > 0.0001), "quadrant"] <- 3

#mapping clusters
tm_shape(m1_sf) + tm_fill(col = "quadrant", title = "",
breaks = c(1, 2, 3, 4),
palette =  c("red", "blue", "white"),
labels = c("High-High", "Low-Low", "Non-significant")) + 
tm_legend(text.size = 1)  + tm_borders(alpha = 0.5) +
tm_layout(frame = FALSE,  title = "Clusters")  +
tm_layout(legend.outside = TRUE)
```

Getis-Ord's G* - to identify significant clusters
```{r}
tmap_mode("plot")

#identify all neighboring polygons for each polygon in the dataset (including self)
lw_self <- nb2listw(include.self(S.dist), style="W",zero.policy=T) 

# calculate Getis-Ord's G*
RI_local_G <- localG(m1.point$WE, lw_self, alternative = "greater")

RI_local_G_att <- attributes(RI_local_G)

m1_sf$gI <- RI_local_G_att$internals[, "G*i"] # Getis-Ord G
m1_sf$gZ <- RI_local_G_att$internals[, "Z(G*i)"] # z-scores
m1_sf$gp <- RI_local_G_att$internals[, "Pr(z > E(G*i))"] # p-values corresponding to alternative

#designating points to quadrants based on Getis-Ord's G* and corresponding p-values
m1_sf$quadrant_g <- NA
# high-high
m1_sf[m1_sf$gI > 0 & (!is.na(m1_sf$gp) & m1_sf$gp <= 0.0001), "quadrant_g"]<- 1
# low-low
#m1_sf[m1_sf$gI < 0 & (!is.na(m1_sf$gp) & m1_sf$gp <= 0.0001), "quadrant_g"]<- 2
# non-significant
m1_sf[(!is.na(m1_sf$gp) & m1_sf$gp > 0.0001), "quadrant_g"] <- 3

#mapping clusters
tm_shape(m1_sf) + tm_fill(col = "quadrant_g", title = "",
breaks = c(1, 3, 4),
palette =  c("red", "white"),
labels = c("High-High", "Non-significant")) + 
tm_legend(text.size = 1)  + tm_borders(alpha = 0.5) +
tm_layout(frame = FALSE,  title = "Clusters")  +
tm_layout(legend.outside = TRUE)
```

## EXPERT DATA ONLY

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` which generates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces" the function; trace = 2 or higher will be more voluminous

```{r}
comm.point_ex <- phyloregion::points2comm(dat = myg_spiders_expert, res = 0.5, lon = "decimalLongitude", lat = "decimalLatitude", species = "species")
plot(comm.point_ex$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.point_ex <- phyloregion::weighted_endemism(comm.point_ex$comm_dat)
str(Endm.mygalomorph.point_ex)

## Join results back to spatial community data matrix
m1.point_ex <- merge(comm.point_ex$map, data.frame(grids=names(Endm.mygalomorph.point_ex), WE=Endm.mygalomorph.point_ex), by="grids")

# Drop NA in WE
m1.point_ex <- m1.point_ex[!is.na(m1.point_ex$WE),]

m1.point_ex
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.point_ex$corrected_endemism <- m1.point_ex$WE/m1.point_ex$richness
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf_ex <- 
  sf::st_as_sf(m1.point_ex) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = WE)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = corrected_endemism)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf_ex, aes(fill = richness)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#:~:text=Moran's%20I%20values%20significantly%20above,negative%20spatial%20autocorrelation%20or%20dispersion.

```{r}
# extract the center of each polygon
coo_ex <- terra::centroids(m1.point_ex)

coo_ex <- coo_ex |> sf::st_as_sf(crs = 4326)

# Search radius to include all neighboring polygon (0 - 200km)
S.dist_ex  <-  dnearneigh(coo_ex, 0, 200)  

#identify all neighboring polygons for each polygon in the dataset.
lw_ex <- nb2listw(S.dist_ex, style="W",zero.policy=T) 

# Run the MC simulation
R_MI_point_ex  <-  moran.mc(m1.point_ex$richness, lw_ex, nsim=9999,zero.policy=T)
WE_MI_point_ex  <-  moran.mc(m1.point_ex$WE, lw_ex, nsim=9999,zero.policy=T)
CWE_MI_point_ex <-  moran.mc(m1.point_ex$corrected_endemism, lw_ex, nsim=9999,zero.policy=T)

# saveRDS(R_MI_point_ex, "output/analysis/Cleaned_Mygalomorphae_point_R_MI_expert.rds")
# saveRDS(WE_MI_point_ex, "output/analysis/Cleaned_Mygalomorphae_point_WE_MI_expert.rds")
# saveRDS(CWE_MI_point_ex, "output/analysis/Cleaned_Mygalomorphae_point_CWE_MI_expert.rds")
```

Local Moran's I - to identify significant clusters
```{r}
tmap_mode("view")

RI_ex_local <- localmoran(m1.point_ex$corrected_endemism, lw_ex, alternative = "greater")

m1_sf_ex$lmI <- RI_ex_local[, "Ii"] # local Moran's I
m1_sf_ex$lmZ <- RI_ex_local[, "Z.Ii"] # z-scores
m1_sf_ex$lmp <- RI_ex_local[, "Pr(z > E(Ii))"] # p-values corresponding to alternative

# moran's I scatterplot
mp <- moran.plot(as.vector(scale(m1_sf_ex$corrected_endemism)), lw_ex)

#designating points to quadrants based on their position in the moran's scatterplot and the significance of local moran's I
m1_sf_ex$quadrant <- NA
# high-high
m1_sf_ex[(mp$x >= 0 & mp$wx >= 0) & (!is.na(m1_sf_ex$lmp) & m1_sf_ex$lmp <= 0.0001), "quadrant"]<- 1
# low-low
m1_sf_ex[(mp$x <= 0 & mp$wx <= 0) & (!is.na(m1_sf_ex$lmp) & m1_sf_ex$lmp <= 0.0001), "quadrant"]<- 2
# non-significant
m1_sf_ex[(!is.na(m1_sf_ex$lmp) & m1_sf_ex$lmp > 0.0001), "quadrant"] <- 3

#mapping clusters
tm_shape(m1_sf_ex) + tm_fill(col = "quadrant", title = "",
breaks = c(1, 2, 3, 4),
palette =  c("red", "blue", "white"),
labels = c("High-High", "Low-Low", "Non-significant")) +
tm_legend(text.size = 1)  + tm_borders(alpha = 0.5) +
tm_layout(frame = FALSE,  title = "Clusters")  +
tm_layout(legend.outside = TRUE)
```

Getis-Ord's G* - to identify significant clusters
```{r}
tmap_mode("plot")

#identify all neighboring polygons for each polygon in the dataset (including self)
lw_ex_self <- nb2listw(include.self(S.dist_ex), style="W",zero.policy=T) 

# calculate Getis-Ord's G*
RI_local_G_ex <- localG(m1.point_ex$richness, lw_ex_self, alternative = "greater")

RI_local_G_ex_att <- attributes(RI_local_G_ex)

m1_sf_ex$gI <- RI_local_G_ex_att$internals[, "G*i"] # Getis-Ord G
m1_sf_ex$gZ <- RI_local_G_ex_att$internals[, "Z(G*i)"] # z-scores
m1_sf_ex$gp <- RI_local_G_ex_att$internals[, "Pr(z > E(G*i))"] # p-values corresponding to alternative

#designating points to quadrants based on Getis-Ord's G* and corresponding p-values
m1_sf_ex$quadrant_g <- NA
# high-high
m1_sf_ex[m1_sf_ex$gI > 0 & (!is.na(m1_sf_ex$gp) & m1_sf_ex$gp <= 0.0001), "quadrant_g"]<- 1
# low-low
m1_sf_ex[m1_sf_ex$gI < 0 & (!is.na(m1_sf_ex$gp) & m1_sf_ex$gp <= 0.0001), "quadrant_g"]<- 2
# non-significant
m1_sf_ex[(!is.na(m1_sf_ex$gp) & m1_sf_ex$gp > 0.0001), "quadrant_g"] <- 3

#mapping clusters
tm_shape(m1_sf_ex) + tm_fill(col = "quadrant_g", title = "",
breaks = c(1, 2, 3, 4),
palette =  c("red", "blue", "white"),
labels = c("High-High", "Low-Low", "Non-significant")) + 
tm_legend(text.size = 1)  + tm_borders(alpha = 0.5) +
tm_layout(frame = FALSE,  title = "Clusters")  +
tm_layout(legend.outside = TRUE)
```

