---
title: "Endemism analysis"
author: "Fonti Kar, Payal Bal, Aaron Greenfield"
date: "2023-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(phyloregion, terra, sf, sp, tidyverse, ggplot2, ozmaps, tmap, viridis)
```

### Load alpha hull spatial polygons

These are estimated from {ConR}, this is a massive list, each element is a spatial polygon dataframe

```{r}
# species_polys <- readRDS("output/myg_spiders_conr_sp.rds")

species_polys <- readRDS("output/spatial/Spatial_alpha_hulls_cleaned_Mygalomorphae_withassertions_2024-03-13_ALA.rds")
```

### Pre-processing

```{r}
# Combine into one large spatial dataframe
combinedShp <- do.call(what = rbind, args=species_polys)

## Assign species names
combinedShp <- terra::vect(combinedShp)
names(combinedShp) <- "species"

combinedShp
```

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` which generates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces" the function; trace = 2 or higher will be more voluminous

```{r}
comm.poly <- phyloregion::polys2comm(dat = combinedShp, species = "species", trace=1, res = 0.5)
str(comm.poly)
plot(comm.poly$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.poly <- phyloregion::weighted_endemism(comm.poly$comm_dat)
str(Endm.mygalomorph.poly)

## Join results back to spatial community data matrix
m1.poly <- merge(comm.poly$map, data.frame(grids=names(Endm.mygalomorph.poly), WE=Endm.mygalomorph.poly), by="grids")

# Drop NA in WE
m1.poly <- m1.poly[!is.na(m1.poly$WE),]

m1.poly
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.poly$corrected_endemism <- m1.poly$WE/m1.poly$richness
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf <- 
  sf::st_as_sf(m1.poly) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf, aes(fill = corrected_endemism)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot() +
  geom_sf(data = m1_sf, aes(fill = richness)) + 
  geom_sf(data = ozmap_states, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  viridis::scale_fill_viridis(option = "G",direction = -1) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

https://www.paulamoraga.com/book-spatial/spatial-autocorrelation.html#:~:text=Moran's%20I%20values%20significantly%20above,negative%20spatial%20autocorrelation%20or%20dispersion.

```{r}
library(spdep)

# extract the center of each polygon
coo <- terra::centroids(m1.poly)

coo <- coo |> sf::st_as_sf(crs = 4326)

# Search radius to include all neighboring polygon (0 - 200km)
S.dist  <-  dnearneigh(coo, 0, 200000)  

#identify all neighboring polygons for each polygon in the dataset.
lw <- nb2listw(S.dist, style="W",zero.policy=T) 

# Run the MC simulation
RI  <-  moran.mc(m1.poly$richness, lw, nsim=9999,zero.policy=T)
MI  <-  moran.mc(m1.poly$corrected_endemism, lw, nsim=9999,zero.policy=T)
WE_MI  <-  moran.mc(m1.poly$WE, lw, nsim=9999,zero.policy=T)

saveRDS(RI, "output/analysis/Cleaned_Mygalomorphae_RI.rds")
saveRDS(MI, "output/analysis/Cleaned_Mygalomorphae_MI.rds")
saveRDS(WE_MI, "output/analysis/Cleaned_Mygalomorphae_WE_MI.rds")
```

