---
title: "Endemism analysis"
author: "Fonti Kar, Payal Bal, Aaron Greenfield"
date: "2023-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(phyloregion, terra, sf, sp, tidyverse, ggplot2, ozmaps)
```

### Load alpha hull spatial polygons

These are estimated from {ConR}, this is a massive list, each element is a spatial polygon dataframe

```{r}
species_polys <- readRDS("output/myg_spiders_conr_sp.rds")
```

### Pre-processing

```{r}
# Combine into one large spatial dataframe
combinedShp <- do.call(what = rbind, args=species_polys)

## Assign species names
combinedShp$dummy <- names(species_polys)
combinedShp <- terra::vect(combinedShp)
names(combinedShp) <- "species"

combinedShp
```

### Calculate Community Matrix

Convert raw input distribution data to community at 0.5 decimal degrees
`trace = 1` is passed to `mean_dist()` whichgenerates mean pairwise distance matrix from a set many pairwise distance matrices. Note: all matrices should be of the same dimension. `trace` "traces"the function; trace = 2 or higher will be more voluminous

```{r}
comm.poly <- phyloregion::polys2comm(dat = combinedShp, species = "species", trace=1, res = 0.5)
str(comm.poly)
plot(comm.poly$map)
```

### Calulate weighed endemism 

```{r}
Endm.mygalomorph.poly <- phyloregion::weighted_endemism(comm.poly$comm_dat)
str(Endm.mygalomorph.poly)

## Join results back to spatial community data matrix
m1.poly <- merge(comm.poly$map, data.frame(grids=names(Endm.mygalomorph.poly), WE=Endm.mygalomorph.poly), by="grids")

# Drop NA in WE
m1.poly <- m1.poly[!is.na(m1.poly$WE),]

m1.poly
```


### Calulate corrected weighted endemism

Weighted endemism tally per cell divided by the species richness of that cell

```{r}
m1.poly$corrected_endemism <- m1.poly$WE/m1.poly$richness

plot(m1.poly)
```

### Create a endemism map

Convert to `sf` 

```{r}
m1_sf <- 
  sf::st_as_sf(m1.poly) |> 
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = m1_sf, fill = NA, colour = "black") + 
  geom_sf(data = ozmap_country, fill = NA, colour = "black") + 
  coord_sf(xlim = c(110, 155), 
           ylim = c(-10, -45)) + 
  theme_minimal()
```

### Reproducing Aaron's original maps

Species richness

```{r}
library(tmap)

aust_WGS84 <- st_transform(ozmaps::ozmap_country, crs(comm.poly$map))

# Adjusting the axes
bbox_new <- st_bbox(m1_sf)
xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin

bbox_new[1] <- bbox_new[1] + (0.62 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] - (0.1 * xrange) # xmax - right
bbox_new[2] <- bbox_new[2] - (0.01 * yrange) # ymin - bottom
bbox_new[4] <- bbox_new[4] - (0.1 * yrange) # ymax - top

spRh.poly <- tm_shape(m1_sf, 
                      bbox = bbox_new
                      ) +
  tm_polygons("richness", 
              style="quantile", 
              title="Species Richness",
              palette="YlGnBu") +
  tm_shape(aust_WGS84) +
  tm_polygons("NAME", 
              alpha = 0,
              legend.show = F,
              lwd = 2, border.col = 1)

spRh.poly
```

```{r}
weighted.endemism.poly <- tm_shape(m1_sf, bbox = bbox_new) +
  tm_polygons("WE", 
              style="quantile", 
              title="Weighted endemism",
              palette="YlGnBu")+
  tm_shape(aust_WGS84)+
  tm_polygons("NAME", 
              alpha = 0,
              legend.show = F,
              lwd = 2, border.col = 1)

weighted.endemism.poly
```


```{r}
Corrected.weighted.endemism.poly <- tm_shape(m1_sf, bbox = bbox_new) +
  tm_polygons("corrected_endemism", 
              style="quantile", 
              title="Polygons: Corrected weighted endemism",
              lwd = 0.5,
              palette="YlGnBu")+
  tm_shape(aust_WGS84)+
  tm_polygons("NAME", 
              alpha = 0,
              legend.show = F,
              lwd = 2, border.col = 1)

Corrected.weighted.endemism.poly
```

