---
title: "Data cleaning"
author: "Fonti Kar"
date: "2023-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cleaning data from ALA

This document lays out the workflow of cleaning data for Mygalomorph spiders from the ALA database 

## Load dependencies

```{r loadpackages}
# install.packages("pacman")
pacman::p_load(galah, arrow, here, tidyverse)
```

## Read in parquet

```{r}
myg_spiders_all <- open_dataset(here("data/galah/Mygalomorphae_withassertions2023-09-01_ALA.paraquet"))
```

## Subset columns we want

```{r}
myg_spiders <- myg_spiders_all |> 
  select(decimalLatitude:collectionCode) |> 
  collect()
```

### Data Overview

How many species/genus in dataset

```{r}
myg_spiders |> 
  group_by(taxonRank) |> 
  summarise(n = n())

# Why are these unranked? Selenocosmia crassipes AB to ask colleagues and JM 
myg_spiders |> 
  filter(taxonRank == "unranked") |> 
  select(scientificName, collectionCode) |> 
  distinct()
```

Across states

<!-- AB to do this -->

```{r}

```

General map of occurrences

```{r}

```

### Geographic errors

<!-- Clean for geographic errors: missing or invalid lat/long; if data needs to be masked by land/water. There are also R packages like coordinate cleaner and others (you'd know more) to clean occurrence data  -->

```{r}

```

### Taxonomic errors

<!-- Clean for taxonomic errors: misspelt names, incorrect, synonyms.  -->
Excluding introduced/invasives taxa
Excluding marines

```{r}

```


<!-- Notes from Payal -->
<!-- Decide whether to keep morpho-species  -->
<!-- How do you distinguish these in ALA data? -->
<!-- FK to check with PB, I don't think these in are ALA data -->
```{r}

```

### Subspecies

<!-- Decide whether to keep sub-species and if not whether to combine their data with the parent species -->
 
```{r}

```


<!-- Check names against a naming authority to make sure all species included in the analyses are valid -->

<!-- Check for duplicates (as the last step) -->

In this step we can inspect and clean the data of duplicates 

```{r}
nrow(Mygalomorphae_occurrences)
head(Mygalomorphae_occurrences)

Mygalomorphae_clean <- Mygalomorphae_occurrences |> 
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) |>
  filter(!duplicated(decimalLatitude) & !duplicated(decimalLongitude)) 

nrow(Mygalomorphae_clean)
head(Mygalomorphae_clean)
```

